---
title: "Accounting: from a raw ledger to net dollars per year"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accounting: from a raw ledger to net dollars per year}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
ex <- us_sample_extract()
```

`us_panel()` runs a fixed sequence of accounting rules between the raw
transaction ledger and the org × award × year panel. This vignette walks the
rules in order, with the measured pilot evidence behind each. (The full
specification lives in the repository's `ACCOUNTING.md`.)

## 1. Clean the ledger — `us_normalize_transactions()`

```{r}
tx <- us_normalize_transactions(ex$transactions)
```

Five rules, each grounded in the 50-org pilot (302,025 rows):

1. **De-duplicate on the transaction key.** Download batches overlap and one
   organization's several UEIs can appear in multiple jobs. The pilot carried
   6.9% exact duplicates; removing them recovered 5,754 awards that had been
   failing the lifetime reconciliation.
2. **Corrections are kept; deletes are dropped.** Assistance rows carry
   `correction_delete_indicator_code`: `D` withdraws a record; `C` marks the
   row as *being* the current correction. **37.6% of pilot assistance rows
   carry `C`** — a pipeline that misreads `C` as "superseded" throws away a
   third of the ledger.
3. **Drop county-level aggregate records** (assistance `record_type_code`
   1) — they have no identified recipient and double-count entity-level money.
4. **Keep zero-dollar actions, flagged.** They carry period-of-performance
   changes and they are how activity is counted.
5. **Flag anomalies, never silently repair** — a non-zero amount on an
   administrative action, dates outside the period of performance, a
   recipient UEI that was never requested.

## 2. The sign is the truth

Action-type labels describe *intent*; the amount's sign describes what
happened. The sample shows both mismatches at once — a `CONTINUATION`
carrying $0 and a `REVISION` carrying a claw-back:

```{r}
tx[award_key %in% c("ASST_NON_HDTRA12310001_097", "ASST_NON_D8732143_075") &
   action_type_label %in% c("Continuation", "Revision"),
   .(award_key, action_date, action_type_label, federal_action_obligation)][order(award_key, action_date)]
```

In the pilot, 8.6% of all transactions are negative, and de-obligations
arrive under **every** action-type label. So the ledger (`us_ledger()`)
classifies money by sign and keeps `action_class` as context — and negatives
are *kept*: dropping them inflates every affected year.

## 3. Net by year, keeping gross and net — `us_net_by_year()`

Every org-award-year carries `obligation_positive`, `obligation_negative`,
and `obligation_net`. A cell netting to $1M from +$5M and −$4M is a
different fact from a clean $1M, and the panel never carries the net alone:

```{r, warning = FALSE}
p <- us_panel(ex)
p$panel[award_key == "ASST_NON_HDTRA12310001_097",
        .(year, obligation_positive, obligation_negative, obligation_net)]
```

## 4. Which year does a claw-back belong to?

The hard question. The sample's claw-back — $310,999 reversed in 2024
against money obligated years earlier — can be booked two defensible ways,
and the package implements both rather than choosing silently:

| policy | books the reversal in | character |
|---|---|---|
| `as_posted` (default) | the year it happened | cash-like; matches USAspending's own presentation; each year reproducible from that year's transactions |
| `restate` | the year(s) that carried the money, latest-first | accrual-like; but rewrites history as new claw-backs arrive |
| `drop` | nowhere | diagnostic only — measures the overstatement; warns |

```{r, warning = FALSE}
lg <- us_ledger(tx)
om <- us_org_map(ex$meta$uei)
for (pol in c("as_posted", "restate")) {
  g <- us_net_by_year(lg, om, deobligation_policy = pol)
  cat(pol, ":\n")
  print(g[award_key == "ASST_NON_D8732143_075", .(year, obligation_net)])
}
```

Here the two agree — the obligation being reversed predates this extract's
window, so `restate` has nothing in-window to push the reversal onto and
correctly leaves it where it was posted. **Measured on the pilot**
($118.5bn, 2007–2025): `restate` moves most years by under 1%; the policies
genuinely diverge only at the data's leading edge (2024 −$460m, 2025
+$646m), which is where the numbers are still moving anyway. `as_posted` is
therefore a safe default for historical work. Under it, about 10% of pilot
org-award-years net negative — real, and kept.

## 5. Year basis

`period = "calendar"` (default) books on the calendar year of `action_date`;
`"fiscal"` uses the federal fiscal year (starting 1 October). Both are
recomputed from `action_date` rather than trusting the source's fiscal-year
column, so the two bases are guaranteed consistent with each other and with
the dates.

## 6. What the panel then is

`obligation_net` per org-award-year, minus pass-through where fetched
(`vignette("subawards")`), with every rule above checkable after the fact:
`us_reconcile()` tests each award's summed actions against USAspending's
independently computed lifetime total (`vignette("reconciliation")`).
