---
title: "Accounting: from a raw ledger to net dollars per year"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accounting: from a raw ledger to net dollars per year}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
ex <- us_sample_extract()
```

`us_panel()` runs a fixed sequence of accounting rules between the raw
transaction ledger and the org × award × year panel. This vignette walks the
rules in order, with the measured pilot evidence behind each. (The full
specification lives in the repository's `ACCOUNTING.md`.)

## 1. Clean the ledger — `us_normalize_transactions()`

```{r}
tx <- us_normalize_transactions(ex$transactions)
```

Five rules, each grounded in the 50-org pilot (302,025 rows):

1. **De-duplicate on the transaction key.** Download batches overlap and one
   organization's several UEIs can appear in multiple jobs. The pilot carried
   6.9% exact duplicates; removing them recovered 5,754 awards that had been
   failing the lifetime reconciliation.
2. **Corrections are kept; deletes are dropped.** Assistance rows carry
   `correction_delete_indicator_code`: `D` withdraws a record; `C` marks the
   row as *being* the current correction. **37.6% of pilot assistance rows
   carry `C`** — a pipeline that misreads `C` as "superseded" throws away a
   third of the ledger.
3. **Drop county-level aggregate records** (assistance `record_type_code`
   1) — they have no identified recipient and double-count entity-level money.
4. **Keep zero-dollar actions, flagged.** They carry period-of-performance
   changes and they are how activity is counted.
5. **Flag anomalies, never silently repair** — a non-zero amount on an
   administrative action, dates outside the period of performance, a
   recipient UEI that was never requested.

## 2. The sign is the truth

Action-type labels describe *intent*; the amount's sign describes what
happened. The sample shows both mismatches at once — a `CONTINUATION`
carrying $0 and a `REVISION` carrying a claw-back:

```{r}
tx[award_key %in% c("ASST_NON_HDTRA12310001_097", "ASST_NON_D8732143_075") &
   action_type_label %in% c("Continuation", "Revision"),
   .(award_key, action_date, action_type_label, federal_action_obligation)][order(award_key, action_date)]
```

In the pilot, 8.6% of all transactions are negative, and de-obligations
arrive under **every** action-type label. So the ledger (`us_ledger()`)
classifies money by sign and keeps `action_class` as context — and negatives
are *kept*: dropping them inflates every affected year.

## 3. Net by year, keeping gross and net — `us_net_by_year()`

Every org-award-year carries `obligation_positive`, `obligation_negative`,
and `obligation_net`. A cell netting to $1M from +$5M and −$4M is a
different fact from a clean $1M, and the panel never carries the net alone:

```{r, warning = FALSE}
p <- us_panel(ex)
p$panel[award_key == "ASST_NON_HDTRA12310001_097",
        .(year, obligation_positive, obligation_negative, obligation_net)]
```

## 4. Which year does a claw-back belong to?

The hard question. The sample's claw-back — $310,999 reversed in 2024
against money obligated years earlier — can be booked two defensible ways,
and the package implements both rather than choosing silently:

| policy | books the reversal in | character |
|---|---|---|
| `as_posted` (default) | the year it happened | cash-like; matches USAspending's own presentation; each year reproducible from that year's transactions |
| `restate` | the year(s) that carried the money, latest-first | accrual-like; but rewrites history as new claw-backs arrive |
| `drop` | nowhere | diagnostic only — measures the overstatement; warns |

```{r, warning = FALSE}
lg <- us_ledger(tx)
om <- us_org_map(ex$meta$uei)
for (pol in c("as_posted", "restate")) {
  g <- us_net_by_year(lg, om, deobligation_policy = pol)
  cat(pol, ":\n")
  print(g[award_key == "ASST_NON_D8732143_075", .(year, obligation_net)])
}
```

Here the two agree — the obligation being reversed predates this extract's
window, so `restate` has nothing in-window to push the reversal onto and
correctly leaves it where it was posted. **Measured on the pilot**
($118.5bn, 2007–2025): `restate` moves most years by under 1%; the policies
genuinely diverge only at the data's leading edge (2024 −$460m, 2025
+$646m), which is where the numbers are still moving anyway. `as_posted` is
therefore a safe default for historical work. Under it, about 10% of pilot
org-award-years net negative — real, and kept.

## 5. Year basis

`period = "calendar"` (default) books on the calendar year of `action_date`;
`"fiscal"` uses the federal fiscal year (starting 1 October). Both are
recomputed from `action_date` rather than trusting the source's fiscal-year
column, so the two bases are guaranteed consistent with each other and with
the dates.

## 6. What the panel then is

`obligation_net` per org-award-year, minus pass-through where fetched
(`vignette("subawards")`), with every rule above checkable after the fact:
`us_reconcile()` tests each award's summed actions against USAspending's
independently computed lifetime total (`vignette("reconciliation")`).

## 7. Inside the object `us_panel()` returns

`us_panel()` returns a list of class `usaspend_panel` holding six elements.
The first five are tables; `meta` records the choices the build was made
under (year basis, money measure, de-obligation policy, whether outbound
subawards were fetched).

| element | grain | what it is |
|---|---|---|
| `panel` | org × award × year | **the deliverable** — net dollars per award per year |
| `awards` | award | the spine: one row per award with its attributes and lifetime totals |
| `transactions` | award action | the normalized ledger the panel was aggregated from |
| `subawards` | FSRS report line | subaward rows with `direction` assigned (`in`/`out`/`internal`/`unrelated`) |
| `subawards_in` | org × year | subaward revenue *received* — money invisible in prime data |
| `meta` | — | build parameters and provenance |

The previews below use the packaged Vanderbilt University Medical Center
data (`?vumc_transactions`), rebuilt into a full panel object:

```{r vumc-build, message = FALSE, warning = FALSE}
library(data.table)
data(vumc_transactions); data(vumc_subawards)
ueis <- unique(vumc_transactions$recipient_uei)

ex_v <- structure(
  list(transactions = vumc_transactions, subawards = vumc_subawards,
       jobs = NULL,
       meta = list(uei = ueis, years = 2008:2025, source = "api",
                   subawards = "both", extracted_at = Sys.time())),
  class = "usaspend_extract")

p <- us_panel(ex_v, org_map = data.frame(uei = ueis, org_id = "35-2528741"),
              fill_gaps = TRUE)
```

```{r kable-helper, include = FALSE}
show10 <- function(d, height = "300px") {
  kableExtra::kbl(head(d, 10), format = "html", row.names = FALSE) |>
    kableExtra::kable_styling(font_size = 11, bootstrap_options =
                                c("striped", "condensed")) |>
    kableExtra::scroll_box(width = "100%", height = height)
}
dict <- function(...) {
  x <- matrix(c(...), ncol = 2, byrow = TRUE)
  kableExtra::kbl(data.frame(field = x[, 1], meaning = x[, 2]),
                  format = "html", row.names = FALSE) |>
    kableExtra::kable_styling(font_size = 11, bootstrap_options =
                                c("striped", "condensed")) |>
    kableExtra::scroll_box(width = "100%", height = "260px")
}
```

### 7.1 `panel` — org × award × year

```{r}
dim(p$panel)
```

```{r echo = FALSE}
dict(
  "org_id",                  "Organization id from your crosswalk (EIN here); the unit of analysis.",
  "award_key",               "Generated unique award id (stable, unlike FAIN/PIID).",
  "year / year_basis",       "Calendar or federal fiscal year, recomputed from action_date.",
  "award_group / award_family / award_type_code / award_type_label", "Contract vs assistance, and the finer type.",
  "awarding_agency_code / awarding_agency_name / awarding_sub_agency_name / funding_agency_name", "Who awarded and who funded.",
  "cfda_number / naics_code / recipient_state", "Program, industry, registered state.",
  "n_transactions / n_actions_positive / n_actions_negative", "Activity counts. n_transactions = 0 marks a fill-gaps placeholder row.",
  "obligation_positive / obligation_negative / obligation_net", "Gross new money, gross claw-backs, and their sum. Never carried as net alone.",
  "deobligation_prior_year", "Reversals pushed back by the restate policy (zero under as_posted).",
  "loan_face_value / loan_subsidy_cost", "Loan amounts; never revenue.",
  "subaward_out_amount / n_subawards_out", "Pass-through paid onward under this award-year.",
  "subaward_in_amount / n_subawards_in", "Subawards received under this award (rare at award grain; see subawards_in).",
  "net_revenue",             "obligation_net - subaward_out_amount.",
  "flags",                   "Anomaly markers carried from normalization.")
```

```{r echo = FALSE}
show10(p$panel[order(-obligation_net)])
```

### 7.2 `awards` — the spine

One row per award, each attribute taken from the latest action that reports
it (award-level fields are sparse across modifications).

```{r}
dim(p$awards)
```

```{r echo = FALSE}
dict(
  "award_key / award_id / parent_award_id", "Generated key, the human FAIN/PIID, and the parent IDV where present.",
  "award_group / award_family / award_type_code / award_type_label", "What kind of award.",
  "recipient_uei / recipient_name / recipient_parent_uei / recipient_state", "Recipient as of the latest action.",
  "awarding_agency_code / awarding_agency_name / awarding_sub_agency_name / funding_agency_name", "Agencies, as of the latest action.",
  "cfda_number / cfda_title / naics_code / psc_code", "Program and industry classifiers.",
  "base_action_date / latest_action_date", "First and last observed actions.",
  "pop_start_date / pop_end_date", "Period of performance.",
  "total_obligated / total_outlayed", "USAspending's own lifetime totals — the reconciliation reference (outlays mostly missing pre-2017).",
  "potential_value", "Ceiling, not money; never summed as revenue.",
  "subaward_count / subaward_total", "From the award overview endpoint when enriched.",
  "n_transactions / n_years / n_recipients", "Ledger footprint. n_recipients > 1 marks awards that moved between institutions.")
```

```{r echo = FALSE}
show10(p$awards[order(-total_obligated)])
```

### 7.3 `transactions` — the normalized ledger

Same schema as `vumc_transactions` (see `?vumc_transactions` for the full
field dictionary) plus the normalization columns:

```{r echo = FALSE}
dict(
  "is_zero_dollar",  "TRUE on $0 actions — kept, because they carry period-of-performance changes and activity counts.",
  "is_deobligation", "TRUE where the obligation is negative.",
  "is_legacy_cdi",   "Undocumented correction/delete codes (kept, flagged).",
  "is_stray_uei",    "Recipient UEI was never requested (API text-matching stray).",
  "flags",           "Semicolon-joined anomaly markers (money on an admin action, action after period of performance, ...).")
```

```{r echo = FALSE}
show10(p$transactions[order(federal_action_obligation)])
```

### 7.4 `subawards` — with direction

```{r}
p$subawards[, .(rows = .N,
                dollars_m = round(sum(subaward_amount, na.rm = TRUE) / 1e6)),
            by = direction]
```

`in` rows are revenue (VUMC as subawardee); `out` rows are pass-through
(VUMC as prime); `internal` rows move money between an organization's own
UEIs and are excluded from both flows; `unrelated` rows sit on
multi-recipient awards whose current holder is someone else.

```{r echo = FALSE}
show10(p$subawards[direction == "out"][order(-subaward_amount)])
```

### 7.5 `subawards_in` — inbound revenue by org-year

Money received as a subrecipient never appears in prime award data — an
organization funded purely this way looks like a non-recipient. It is
therefore kept as its own org × year table rather than being forced onto
award rows.

```{r echo = FALSE}
show10(p$subawards_in[order(-subaward_in_amount)], height = "220px")
```

### 7.6 `meta`

```{r}
str(p$meta)
```

## 8. One organization end to end: Vanderbilt University Medical Center

VUMC (EIN 35-2528741) is the packaged example because it exercises every
rule at once: the highest pass-through share in the 50-org pilot, claw-backs
in every year, and both subaward directions at scale.

### 8.1 The raw ledger

```{r}
nrow(vumc_transactions)
vumc_transactions[, .N,
  by = .(sign = fcase(federal_action_obligation > 0, "positive",
                      federal_action_obligation < 0, "negative",
                      default = "zero"))]
```

Of `r nrow(vumc_transactions)` raw transactions, about a quarter are
zero-dollar administrative actions and
`r vumc_transactions[federal_action_obligation < 0, .N]` are de-obligations —
real claw-backs, arriving under every action-type label. On the subaward
side (after normalization assigns direction):

```{r}
p$subawards[direction %in% c("in", "out"),
            .(rows = .N,
              dollars_m = round(sum(subaward_amount, na.rm = TRUE) / 1e6)),
            by = direction]
```

### 8.2 What a claw-back looks like

Award `ASST_NON_U54CA280915_075` (an NIH cooperative agreement) obligates
across several years and then loses $3.8M in a single action. Note that the
sign carries the truth: the claw-back arrives as an ordinary `REVISION`, not
under any special cancellation type.

```{r echo = FALSE}
show10(vumc_transactions[award_key == "ASST_NON_U54CA280915_075"][
  order(action_date),
  .(modification_number, action_date, action_type_label, action_class,
    federal_action_obligation, award_total_obligated)], height = "220px")
```

Summed, the transactions still reconcile to the award's reported lifetime
total — the claw-back is part of the award's true value, not noise:

```{r}
vumc_transactions[award_key == "ASST_NON_U54CA280915_075",
                  .(tx_sum = sum(federal_action_obligation),
                    reported = unique(award_total_obligated))]
```

### 8.3 How pass-through is structured

Pass-through is not on the transaction rows at all. It arrives as separate
FSRS report lines keyed to the *prime award*, each naming a subawardee, an
amount, and a commitment date. VUMC's largest conduit is
`ASST_NON_NU2GGH001943_075` — a CDC global-health cooperative agreement
under which VUMC passes onward far more than it keeps:

```{r echo = FALSE}
show10(p$subawards[direction == "out" &
                     prime_award_key == "ASST_NON_NU2GGH001943_075"][
  order(-subaward_amount),
  .(subaward_number, subawardee_name, subaward_amount, subaward_action_date,
    subaward_year)], height = "220px")
```

### 8.4 When a placeholder is created

FSRS reports trail the obligations they draw on, so pass-through often lands
in an award-year with no prime activity. With `fill_gaps = TRUE`, such years
get a zero-obligation placeholder row (`n_transactions = 0`) so the dollars
have somewhere to land instead of being dropped:

```{r}
p$panel[n_transactions == 0, .N]                      # placeholder rows
p$panel[n_transactions == 0 & subaward_out_amount > 0][
  order(-subaward_out_amount)][1:3,
  .(award_key, year, obligation_net, subaward_out_amount, net_revenue)]
```

### 8.5 The normalized panel

```{r}
dim(p$panel)
p$panel[, .(awards = uniqueN(award_key)), by = award_group]
```

`r data.table::uniqueN(p$panel$award_key)` awards collapse to
`r nrow(p$panel)` org-award-year rows, funded by
`r data.table::uniqueN(p$panel$awarding_agency_name)` federal agencies:

```{r}
p$panel[, .(awards = uniqueN(award_key),
            net_m = round(sum(obligation_net) / 1e6)),
        by = awarding_agency_name][order(-net_m)]
```

```{r fig-agencies, fig.width = 7.5, fig.height = 3.8, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
ag <- p$panel[, .(net_m = sum(obligation_net) / 1e6),
              by = awarding_agency_name][order(-net_m)]
ggplot(ag, aes(reorder(awarding_agency_name, net_m), net_m)) +
  geom_col(fill = "#1a5276") +
  coord_flip() +
  labs(x = NULL, y = "total net obligations ($M)",
       title = "VUMC net obligations by awarding agency, 2008-2025") +
  theme_minimal(base_size = 10)
```

### 8.6 Six awards over their lives

```{r fig-awards, fig.width = 7.5, fig.height = 4.5, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
top6 <- p$panel[, .(m = sum(obligation_net)),
                by = award_key][order(-m)][1:6, award_key]
d6 <- p$panel[award_key %in% top6]
d6[, award := sub("^ASST_NON_|^CONT_AWD_", "", award_key)]
ggplot(d6, aes(year, obligation_net / 1e6)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_line(colour = "#1a5276", linewidth = 0.7) +
  geom_point(colour = "#1a5276", size = 1.6) +
  facet_wrap(~award, ncol = 3, scales = "free_y") +
  scale_x_continuous(breaks = c(2010, 2015, 2020, 2025)) +
  labs(x = NULL, y = "net obligations ($M)",
       title = "VUMC's six largest awards, year by year",
       subtitle = "Points below zero are claw-backs booked as posted") +
  theme_minimal(base_size = 10)
```

### 8.7 The top 20 awards

```{r fig-top20, fig.width = 7.5, fig.height = 5, echo = FALSE, warning = FALSE}
top20 <- p$panel[, .(m = sum(obligation_net) / 1e6,
                     agency = awarding_agency_name[1]),
                 by = award_key][order(-m)][1:20]
top20[, award := sub("^ASST_NON_|^CONT_AWD_", "", award_key)]
ggplot(top20, aes(reorder(award, m), m, fill = agency)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "lifetime net obligations ($M)", fill = NULL,
       title = "Top 20 VUMC awards by lifetime net obligation") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", legend.text = element_text(size = 7))
```

### 8.8 The trend, and what pass-through does to it

```{r fig-trend, fig.width = 7.5, fig.height = 4, echo = FALSE, warning = FALSE}
tr <- p$panel[, .(`net obligations` = sum(obligation_net) / 1e6,
                  `net revenue (after pass-through)` = sum(net_revenue) / 1e6),
              by = year][order(year)]
trl <- data.table::melt(tr, id.vars = "year")
ggplot(trl[year <= 2025], aes(year, value, colour = variable)) +
  scale_x_continuous(breaks = seq(2008, 2025, 3)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_colour_manual(values = c("#1a5276", "#c0392b")) +
  labs(x = NULL, y = "$M per year", colour = NULL,
       title = "VUMC federal funding, 2008-2025",
       subtitle = "The gap between the lines is pass-through: money VUMC is obliged to pay onward") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom")
```

The two lines are the whole argument for fetching outbound subawards: by
obligations alone, VUMC looks roughly twice as large as the revenue it
actually retains. The finished versions of these tables ship in the package:
`data(vumc_transactions)`, `data(vumc_subawards)`, `data(vumc_panel)`.
