---
title: "Choosing an acquisition path: API jobs or annual archives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an acquisition path: API jobs or annual archives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE}
library(usaspend)
```

`us_extract()` is the single entry point for acquisition, and it can fetch
the same data two very different ways:

```{r, eval = FALSE}
# one organization -- the API path
ex <- us_extract("CFFMYPABYAG3", years = 2015:2025)

# thousands of organizations -- the archive path
ex <- us_extract(ueis, years = 2008:2025, source = "archive")
```

Both paths land on the same canonical tables, so nothing downstream knows or
cares which was used. But their costs scale on different axes, and picking
wrong can turn minutes into days. This vignette explains the mechanics of
each, and how `us_extract_plan()` chooses between them.

Every constraint quoted below was **measured live** against
`api.usaspending.gov`, not read off the documentation.

## Path A — API download jobs: cost scales with the number of UEIs

`us_extract(source = "api")` submits bulk-download jobs to
`POST /api/v2/download/transactions/`, polls until the server has built the
files, fetches and unzips them, and harmonizes the CSVs. Each job covers a
small batch of UEIs and takes 9–85 seconds server-side.

What the measurements say about how to drive it:

- **`recipient_search_text` accepts about 20 values per job.** 20 succeeds,
  25 fails — the cap is undocumented. In practice large recipients time out
  server-side well below it, so the package's default batch is **5**, and a
  failed batch is retried one UEI at a time so a single oversized recipient
  cannot poison its whole batch.
- **The job state machine is `ready → running → finished | failed`**, and
  `ready` means *queued*, not done. (Treating anything other than `running`
  as terminal — a natural mistake — reads a freshly-queued job as complete
  and throws the download away.)
- **The search endpoint (`spending_by_award`) silently caps at 10,000
  records.** Extraction always uses the download endpoint, which has no cap.
- **Sustained calls get rate-limited.** The client throttles to 2 requests/s
  with exponential backoff, and failures are recorded explicitly in the job
  manifest — a swallowed error looks exactly like a recipient with no awards.
- **Search is floored at 2007-10-01.** Earlier fiscal years are not reachable
  on either path.

The by-product that matters: a recipient-filtered download also returns
subaward rows **where the queried UEI is the subawardee** — inbound revenue
you cannot get from the archives at all.

## Path B — annual archives: cost is fixed in the number of fiscal years

`us_extract(source = "archive")` downloads USAspending's whole-fiscal-year
Award Data Archive files and filters them locally with duckdb:

```{r, eval = FALSE}
man <- us_archive_manifest(2008:2025) |> us_archive_download()
tx  <- us_archive_filter(ueis, man$csv_dir[1], group = "assistance")
```

Two archives exist per fiscal year — `assistance` and `contracts` (direct
payments, loans, and "other" are folded into assistance). A full-year
assistance archive is ~1.4 GB compressed and unpacks to several CSV parts;
the duckdb scan then filters on `recipient_uei` at roughly a gigabyte a
minute. The whole-history cost is a few hours of downloading and scanning —
**whether you want 100 UEIs or 100,000**.

Two things the archives do not contain:

- **Subawards, in either direction.** The archive files are prime
  transactions only.
- Nothing before FY2008 (the same floor as the API).

## Do the two paths return the same data? Measured: yes, with three caveats

Because the paths draw from the same database through different generators,
the package's whole design rests on them being interchangeable. That was
tested directly: real FY2015 and FY2024 archives (generated 2026-08-06) were
filtered to 130 pilot UEIs and compared row-by-row against the same-window
slice of a same-month API pull (2026-08-27).

**Where it matters, they are identical.**

| cell | rows (API vs archive) | one-sided keys | obligation difference |
|---|---|---|---|
| FY2015 assistance | 8,795 = 8,795 | 0 | \$0 |
| FY2015 contracts | 2,947 = 2,947 | 0 | \$0 |
| FY2024 assistance | 17,225 vs 17,224 | 1 | −\$4,125.69 |
| FY2024 contracts | 2,487 = 2,487 | 0 | \$0 |

Settled years match key-for-key and dollar-for-dollar. Archive files
partition time exactly like an API pull filtered on `action_date`: every row
in an FY file has an action date inside that fiscal year. All mapped column
names are present in both.

**Caveat one: archive contract files transpose two column pairs.** In the
archives, `action_type_code` carries the *description* (`FUNDING ONLY
ACTION`) and `action_type` carries the *code* (`C`); likewise
`idv_type_code` holds the mnemonic (`IDC`) and `idv_type` the letter (`B`).
The column names match perfectly, so a name-based check cannot see it — and
before the package handled it, 78% of archive contract actions silently
misclassified. `us_harmonize_transactions()` now detects the transposition
from the value shapes and un-swaps it, so either layout classifies
identically; `us_archive_verify_schema()` warns when it samples a transposed
archive. Assistance files are unaffected.

**Caveat two: award-level derived stamps drift between generation dates.**
Transaction facts (dates, amounts, parties) matched exactly. Two columns did
not, on a handful of rows: `award_total_obligated` (an award-*lifetime*
figure recomputed at file-generation time — the two files were generated
three weeks apart, and the diffs were concentrated in NIH awards, whose
lifetime totals move as awards continue elsewhere), and `award_id_uri`,
whose numeric suffix is a generated identifier that is simply not stable
across systems. Key on `award_key`, never on `award_id_uri`; treat
award-lifetime columns as a property of the *pull date*, not of the
transaction.

**Caveat three: at the leading edge, the archive is up to a month stale.**
The archives regenerate monthly; corrections posted after generation reach
the API first. Measured across the three-week gap, the entire divergence on
FY2024 was **one row** — a NASA de-obligation of −\$4,125.69 modified
fifteen days after the archive was cut — plus re-corrected fields on six of
17,224 common rows: 0.00008% of that year's dollars. Settled years show
zero. If the last few weeks matter for your analysis, refresh the leading
fiscal year over the API (or apply the archive `Delta` files); for anything
historical, the archive is exact.

`us_archive_verify_schema()` runs the name check and the transposition probe
against any unpacked archive in one call — still worth running on a fiscal
year the package has not seen before.

## The crossover

API cost grows with UEIs; archive cost is flat. They cross somewhere in the
low thousands of UEIs. `us_extract_plan()` computes the comparison from the
measured constants — it makes no network calls, so you can plan offline:

```{r}
# a batch of 40 organizations
us_extract_plan(sprintf("DEMO%08d", 1:40), years = 2008:2025)
```

```{r}
# the same question at 5,000 organizations
us_extract_plan(sprintf("DEMO%08d", 1:5000), years = 2008:2025)
```

Measured order-of-magnitude timings at 18 fiscal years:

| input | API path | archive path |
|---|---|---|
| 1 org (3 UEIs) | ~1 min | ~4 hours |
| 1,364 UEIs | ~2 hours | ~4 hours |
| 126,784 UEIs | ~6 days | ~4 hours |

`us_extract(source = "auto")` runs the plan and follows its recommendation.

## What neither path gives you: pass-through

Netting out pass-through — computing
`net_revenue = obligation_net − subaward_out_amount` — needs subawards where
your organization is the **prime**. A recipient-filtered API pull never
returns those (it matches on the subawardee), and the archives carry no
subawards at all. Outbound subawards must be fetched per prime award:

```{r, eval = FALSE}
ex <- us_extract(ueis, years = 2008:2025, source = "archive", subawards = "out")
```

`subawards = "out"` walks the extracted award keys through
`us_fetch_subawards_out()` — batched about 400 award ids per request, with a
count-first screen so awards reporting no subawards are skipped. Budget
roughly one extra API call per few hundred awards. Until that pass runs,
`net_revenue` in the panel equals `obligation_net`, and the panel's `flags`
column says so.

Inbound subawards — revenue received as a subrecipient — are a free
by-product on the API path but absent from an archive extract. Passing
`subawards = "in"` (or `"both"`) with `source = "archive"` appends them
through `us_fetch_subawards_in()`, which runs the standard API download jobs
and harvests only the subaward files. There is no cheaper route — a
subaward-only custom download job exists but builds far slower server-side
than the standard job — so this costs what the API path costs, and is only
worth it below the crossover scale.

The full append matrix:

| you extracted via | inbound (`"in"`) | outbound (`"out"`) |
|---|---|---|
| API | already in the extract | `us_fetch_subawards_out()` per award |
| archive | `us_fetch_subawards_in()` — API-job cost | `us_fetch_subawards_out()` per award |

## Choosing, in one table

| situation | use |
|---|---|
| one organization, or a handful | `source = "api"` |
| up to ~2,000 UEIs | `source = "api"` (check `us_extract_plan()`) |
| thousands of UEIs or more | `source = "archive"` |
| you need inbound subawards | the API path (archives have none) |
| you need outbound subawards / true net revenue | either path **plus** `subawards = "out"` |
| unsure | `source = "auto"` |

Both paths cache downloads under `us_cache_dir()`, so re-runs skip completed
work; `us_cache_status()` shows what is already on disk.
