---
title: "Choosing an acquisition path: API jobs or annual archives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an acquisition path: API jobs or annual archives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE}
library(usaspend)
```

`us_extract()` is the single entry point for acquisition, and it can fetch
the same data two very different ways:

```{r, eval = FALSE}
# one organization -- the API path
ex <- us_extract("CFFMYPABYAG3", years = 2015:2025)

# thousands of organizations -- the archive path
ex <- us_extract(ueis, years = 2008:2025, source = "archive")
```

Both paths land on the same canonical tables, so nothing downstream knows or
cares which was used. But their costs scale on different axes, and picking
wrong can turn minutes into days. This vignette explains the mechanics of
each, and how `us_extract_plan()` chooses between them.

Every constraint quoted below was **measured live** against
`api.usaspending.gov`, not read off the documentation.

## Path A — API download jobs: cost scales with the number of UEIs

`us_extract(source = "api")` submits bulk-download jobs to
`POST /api/v2/download/transactions/`, polls until the server has built the
files, fetches and unzips them, and harmonizes the CSVs. Each job covers a
small batch of UEIs and takes 9–85 seconds server-side.

What the measurements say about how to drive it:

- **`recipient_search_text` accepts about 20 values per job.** 20 succeeds,
  25 fails — the cap is undocumented. In practice large recipients time out
  server-side well below it, so the package's default batch is **5**, and a
  failed batch is retried one UEI at a time so a single oversized recipient
  cannot poison its whole batch.
- **The job state machine is `ready → running → finished | failed`**, and
  `ready` means *queued*, not done. (Treating anything other than `running`
  as terminal — a natural mistake — reads a freshly-queued job as complete
  and throws the download away.)
- **The search endpoint (`spending_by_award`) silently caps at 10,000
  records.** Extraction always uses the download endpoint, which has no cap.
- **Sustained calls get rate-limited.** The client throttles to 2 requests/s
  with exponential backoff, and failures are recorded explicitly in the job
  manifest — a swallowed error looks exactly like a recipient with no awards.
- **Search is floored at 2007-10-01.** Earlier fiscal years are not reachable
  on either path.

The by-product that matters: a recipient-filtered download also returns
subaward rows **where the queried UEI is the subawardee** — inbound revenue
you cannot get from the archives at all.

## Path B — annual archives: cost is fixed in the number of fiscal years

`us_extract(source = "archive")` downloads USAspending's whole-fiscal-year
Award Data Archive files and filters them locally with duckdb:

```{r, eval = FALSE}
man <- us_archive_manifest(2008:2025) |> us_archive_download()
tx  <- us_archive_filter(ueis, man$csv_dir[1], group = "assistance")
```

Two archives exist per fiscal year — `assistance` and `contracts` (direct
payments, loans, and "other" are folded into assistance). A full-year
assistance archive is ~1.4 GB compressed and unpacks to several CSV parts;
the duckdb scan then filters on `recipient_uei` at roughly a gigabyte a
minute. The whole-history cost is a few hours of downloading and scanning —
**whether you want 100 UEIs or 100,000**.

Two things the archives do not contain:

- **Subawards, in either direction.** The archive files are prime
  transactions only.
- Nothing before FY2008 (the same floor as the API).

One assumption to verify on first use: the archive CSVs are expected to carry
the same column names as the download endpoint. `us_archive_verify_schema()`
checks that in one call against a real archive and reports exactly which
mapped columns are missing. Run it before trusting your first archive pull.

## The crossover

API cost grows with UEIs; archive cost is flat. They cross somewhere in the
low thousands of UEIs. `us_extract_plan()` computes the comparison from the
measured constants — it makes no network calls, so you can plan offline:

```{r}
# a batch of 40 organizations
us_extract_plan(sprintf("DEMO%08d", 1:40), years = 2008:2025)
```

```{r}
# the same question at 5,000 organizations
us_extract_plan(sprintf("DEMO%08d", 1:5000), years = 2008:2025)
```

Measured order-of-magnitude timings at 18 fiscal years:

| input | API path | archive path |
|---|---|---|
| 1 org (3 UEIs) | ~1 min | ~4 hours |
| 1,364 UEIs | ~2 hours | ~4 hours |
| 126,784 UEIs | ~6 days | ~4 hours |

`us_extract(source = "auto")` runs the plan and follows its recommendation.

## What neither path gives you: pass-through

Netting out pass-through — computing
`net_revenue = obligation_net − subaward_out_amount` — needs subawards where
your organization is the **prime**. A recipient-filtered API pull never
returns those (it matches on the subawardee), and the archives carry no
subawards at all. Outbound subawards must be fetched per prime award:

```{r, eval = FALSE}
ex <- us_extract(ueis, years = 2008:2025, source = "archive", subawards = "out")
```

`subawards = "out"` walks the extracted award keys through
`us_fetch_subawards_out()` — batched about 400 award ids per request, with a
count-first screen so awards reporting no subawards are skipped. Budget
roughly one extra API call per few hundred awards. Until that pass runs,
`net_revenue` in the panel equals `obligation_net`, and the panel's `flags`
column says so.

## Choosing, in one table

| situation | use |
|---|---|
| one organization, or a handful | `source = "api"` |
| up to ~2,000 UEIs | `source = "api"` (check `us_extract_plan()`) |
| thousands of UEIs or more | `source = "archive"` |
| you need inbound subawards | the API path (archives have none) |
| you need outbound subawards / true net revenue | either path **plus** `subawards = "out"` |
| unsure | `source = "auto"` |

Both paths cache downloads under `us_cache_dir()`, so re-runs skip completed
work; `us_cache_status()` shows what is already on disk.
