---
title: "Award types: what is actually in the data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Award types: what is actually in the data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
```

Every transaction carries an award type code, and the code determines
whether its dollars are revenue, capacity, or someone else's money. The
package's taxonomy:

```{r}
us_award_type_codes()
```

Each code maps to an `award_family` (`grant`, `contract`, `idv`,
`direct_payment`, `loan`, `other`) and an `award_group` (`assistance` vs
`contract` — which of USAspending's two file layouts it arrives in).

## What a real portfolio looks like

The 50-nonprofit pilot (302,025 transactions, FY2008–FY2025), by type:

| type | rows | dollars | what it actually is | revenue? |
|---|---|---|---|---|
| Project grants (04) + coop agreements (05) | 212k | $62.8bn | the core | yes |
| Contracts (A–D) | 56.6k | $76.4bn | delivery orders, definitive contracts, POs, BPA calls | yes |
| Direct payments (06, 10) | 19.3k | $3.0bn | almost entirely Pell, work-study, SEOG — student aid moving *through* the institution | separable; usually exclude |
| Direct loans (07) | 6.5k | $0 obligated, **$626bn face value** | Federal Direct Student Loans; the university is a conduit | face value never |
| Formula/block grants (03, 02) | 2.1k | $1.6bn | mostly government-affiliated orgs | yes |
| Other/reimbursable (11) | 1.2k | $0.3bn | DOE/DOD/USDA cooperative R&D | revenue-like |
| IDVs | 4.3k | ~$0.8bn obligated vs **$115bn ceiling** | contract vehicles | obligations only |

Two numbers carry the whole argument: **$626bn of loan face value** and
**$115bn of IDV ceiling**, against **$118.5bn of actual net obligations**.
Summing the wrong dollar column overstates this portfolio roughly sixfold.

## Loans: the data gets it right, keep it that way

Every type-07 row in the pilot is a Federal Direct Student Loan. The loans
are to *students*; the institution originates them. USAspending already
encodes this correctly — `federal_action_obligation` is $0 on every one, and
the face value travels in its own column (`loan_face_value` in the panel).
The rule is simply to never let face value or subsidy cost reach a revenue
total. Direct payments (06, 10) are the same story one step removed —
student aid booked to the school but economically the students' — so they
stay in the ledger with `award_family = "direct_payment"`, and a
nonprofit-revenue analysis usually filters them out explicitly.

## IDVs: vehicles, not money

An indefinite-delivery vehicle (GWAC, IDC, BPA, BOA, FSS) is a contract
*container*. Its ceiling is capacity; the money is obligated on the delivery
orders placed under it — which are separate awards with their own keys. The
sample shows the pair:

```{r}
tx <- us_sample_extract()$transactions
tx[grepl("HHSD200200720032I", award_key),
   .(rows = .N, obligated = sum(federal_action_obligation)),
   by = .(award_key, award_type_label)]
```

The vehicle: $0 across its own modifications, against a $40M potential
value. The delivery orders under it: the actual obligations. Whether to roll
delivery orders up into their parent vehicle is a modelling choice — it
changes award counts and agency attribution, not dollar totals — and the
package defaults to **not** rolling up.

## Two traps the package absorbs so you don't hit them

**Codes collide across families.** `B` is *Continuation* for assistance and
*Supplemental Agreement* for contracts; `C` is *Revision* vs *Funding Only
Action*; `D` is *Adjustment to Completed Award* vs *Change Order*. Any
classification that ignores `award_group` is wrong, which is why
`us_classify_action()` requires it:

```{r}
us_classify_action(c("B", "C"), award_group = "assistance")[, -1]
us_classify_action(c("B", "C"), award_group = "contract")[, -1]
```

**Description strings are not stable — group on codes, never labels.** The
same pull can carry `"DELIVERY ORDER"` and `"DO"` for one code, and archive
files carry different strings again (`vignette("acquisition")`). Package
labels are always derived from the code via `us_classify_award_type()`, so
they are stable no matter what the source file said. Contract IDV rows add
one more wrinkle: their `award_type_code` is blank and the type lives in
`idv_type_code` as a bare letter, which the classifier prefixes (`"B"` →
`"IDV_B"`) because bare `"B"` would read as Purchase Order.
