---
title: "Data dictionary: tables, keys, and relationships"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data dictionary: tables, keys, and relationships}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
```

This is the reference companion to `vignette("data-model")`: what the
package queries at USAspending, what the extract and panel objects contain,
which fields drive the accounting, and what the code values mean. Short
tables here carry only keys and accounting-critical fields; every table
links to its **full dictionary as CSV** in
[`vignettes/data-dictionary/`](https://github.com/Nonprofit-Open-Data-Collective/usaspend/tree/main/vignettes/data-dictionary).

## 1. The source side: what the package actually queries

It is *not* one query per table. The package touches **four server
interfaces**, and they map onto the extract's tables asymmetrically — one
bulk-download job returns both prime transactions *and* subawards; the
archives return transactions only; subawards at scale need their own search
queries.

| # | interface | what one request covers | populates |
|---|---|---|---|
| 1 | `POST /api/v2/download/transactions/` (bulk job) | up to ~20 UEIs × a date range → a zip of 4 CSVs: assistance + contract prime transactions, **plus** FSRS subawards where a queried UEI is the *subawardee* | `transactions` **and** `subawards` (inbound) |
| 2 | Award Data Archive files (`files.usaspending.gov`) | one fiscal year × one family, *all recipients* → filtered locally with duckdb | `transactions` only |
| 3 | `POST /api/v2/search/spending_by_award/` with `subawards = true` | FSRS lines filtered by prime `award_ids` (→ *outbound*) or by `recipient_search_text` (→ *inbound*) | `subawards` |
| 4 | `GET /api/v2/awards/{award_key}/` | one prime award's summary (`subaward_count`, lifetime totals) | the pass-through screen; `awards` enrichment |

Behind those endpoints, USAspending's warehouse — as far as these queries
can see it — looks like this. Two transaction systems (FABS for assistance,
FPDS for contracts) hang off a derived award summary; FSRS subaward reports
reference a prime award and *two* recipients:

```{=html}
<pre class="mermaid" style="text-align: center; background: transparent;">
erDiagram
    RECIPIENT ||--o{ AWARD : "holds (current)"
    AWARD ||--o{ FABS_TRANSACTION : "assistance actions"
    AWARD ||--o{ FPDS_TRANSACTION : "contract actions"
    AWARD ||--o{ FSRS_SUBAWARD : "prime side"
    RECIPIENT ||--o{ FSRS_SUBAWARD : "subawardee side"
    AWARD |o--o{ AWARD : "parent IDV"
    RECIPIENT {
        string uei PK
        string parent_uei FK
        string name
    }
    AWARD {
        string award_key PK "generated_unique_award_id"
        string award_id "FAIN or PIID -- reused, not unique"
        string recipient_uei FK "as of latest action"
        numeric total_obligated "lifetime, recomputed per pull"
    }
    FABS_TRANSACTION {
        string transaction_key PK
        string award_key FK
        date action_date
        numeric federal_action_obligation
        string correction_delete_code
    }
    FPDS_TRANSACTION {
        string transaction_key PK
        string award_key FK
        date action_date
        numeric federal_action_obligation
    }
    FSRS_SUBAWARD {
        string subaward_number "unique per prime only"
        string prime_award_key FK
        string subawardee_uei FK "often null"
        numeric subaward_amount
        date subaward_action_date
    }
</pre>
```

Two source facts worth reading off the diagram: `award_id` (FAIN/PIID) is
**not** unique — agencies reuse it, which is why every join in the package
runs on the generated `award_key`; and an FSRS line has *two* recipient
edges, which is why subaward **direction** exists only relative to an
organization's UEI set.

## 2. The extract object — `us_extract()` / `us_sample_extract()`

Four elements: two data tables plus the acquisition manifest (`jobs`) and
provenance (`meta`). The tables are **harmonized but not normalized** —
duplicates, `D`-flagged deletes, and aggregate records are still present,
which is why `us_panel()` re-normalizes.

```{=html}
<pre class="mermaid" style="text-align: center; background: transparent;">
erDiagram
    AWARD ||--o{ TRANSACTIONS : "award_key (implicit -- spine built later)"
    AWARD |o--o{ SUBAWARDS : "prime_award_key (inbound rows point OUTSIDE the extract)"
    TRANSACTIONS {
        string transaction_key PK
        string award_key FK
        date action_date
        numeric federal_action_obligation
    }
    SUBAWARDS {
        string subaward_key PK "prime|number|date"
        string prime_award_key FK
        string prime_uei
        string subawardee_uei
        numeric subaward_amount
    }
</pre>
```

Note the second edge: for *inbound* subaward rows the prime award belongs to
**another organization**, so `prime_award_key` usually matches nothing in
your own `transactions`. That is by design — inbound rows are revenue that
prime data cannot see.

**`$transactions`** — one row per award action
([full dictionary, 48 fields](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/transactions.csv)):

| field | role in the accounting |
|---|---|
| `transaction_key` | **PK**; the de-duplication key |
| `award_key` | **FK** to the award; every join runs on it |
| `action_date` | the date money is booked on; years recomputed from it |
| `federal_action_obligation` | **the revenue measure**; sign = truth |
| `pragmatic_obligation` | loan-aware variant (subsidy cost for loans) |
| `loan_face_value` / `loan_subsidy_cost` | carried separately; never revenue |
| `award_total_obligated` | award-*lifetime* stamp; the reconciliation target |
| `recipient_uei` | who held the award *at this action* (can change) |
| `award_group` / `award_family` / `award_type_code` | the include/exclude switches |
| `action_type_code` / `action_class` | what kind of action (context only — see §5) |
| `correction_delete_code` | `C` kept, `D` dropped in normalization |
| `record_type_code` | `1` = aggregate record, dropped |
| `last_modified_date` | de-duplication tiebreaker; vintage marker |

**`$subawards`** — one row per FSRS report line
([full dictionary, 24 fields](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/subawards.csv)):

| field | role in the accounting |
|---|---|
| `subaward_key` | **PK**: `prime_award_key \| subaward_number \| action_date`; restatement de-dup key |
| `prime_award_key` | **FK** to the prime award |
| `prime_uei` / `subawardee_uei` | the two sides; direction is computed from them |
| `subaward_amount` | committed dollars (≥ \$30k only; a floor) |
| `subaward_action_date` | the booking date (report dates lag) |
| `report_last_modified` | restatement tiebreaker |
| `direction` | `NA` until `us_normalize_subawards()` assigns it |

**`$jobs`** is the acquisition manifest (download jobs or archive files);
**`$meta`** records `uei`, `years`, `source`, and `subawards` — which
directions were fetched, so a zero can never pass as a measured zero.

## 3. The panel object — `us_panel()`

Five tables plus `meta`. The panel is keyed **org × award × year**;
organization-level inbound revenue lives in its own table because it cannot
sit on another organization's award rows.

```{=html}
<pre class="mermaid" style="text-align: center; background: transparent;">
erDiagram
    ORG ||--o{ PANEL : "org_id"
    AWARDS ||--o{ PANEL : "award_key"
    AWARDS ||--o{ TRANSACTIONS : "award_key"
    AWARDS |o--o{ SUBAWARDS : "prime_award_key (outbound)"
    ORG ||--o{ SUBAWARDS_IN : "org_id x year"
    ORG {
        string org_id PK "from us_org_map()"
    }
    AWARDS {
        string award_key PK
        string recipient_uei "as of latest action"
        numeric total_obligated "reconciliation target"
        int n_recipients "gt 1 = PI transfer"
    }
    TRANSACTIONS {
        string transaction_key PK
        string award_key FK
    }
    PANEL {
        string org_id PK
        string award_key PK
        int year PK
        numeric obligation_net
        numeric subaward_out_amount
        numeric net_revenue
    }
    SUBAWARDS {
        string subaward_key PK
        string prime_award_key FK
        string direction
    }
    SUBAWARDS_IN {
        string org_id PK
        int year PK
        numeric subaward_in_amount
    }
</pre>
```

**`$panel`** — the deliverable
([full dictionary, 30 fields](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/panel.csv)):

| field | role |
|---|---|
| `org_id` + `award_key` + `year` | **composite PK** (grain asserted by `us_audit()`) |
| `obligation_positive` / `obligation_negative` | gross components — never dropped |
| `obligation_net` | their sum; a commitment, not cash; can be negative |
| `subaward_out_amount` | pass-through paid onward |
| **`net_revenue`** | `obligation_net − subaward_out_amount` |
| `loan_face_value` | carried separately, never in `net_revenue` |
| `award_family` | the include/exclude switch (loans, student aid) |
| `flags` | `subawards_out_not_fetched` ⇒ `net_revenue = obligation_net` |

**`$awards`** — the spine
([full dictionary, 33 fields](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/awards.csv)): attributes as of
the latest action, `total_obligated` (the reconciliation target),
`obligated_in_extract` (what this window saw), and `n_recipients` (> 1
flags PI-transfer awards).

**`$transactions` / `$subawards`** — the normalized inputs, same shapes as
§2 plus normalization flags (`is_zero_dollar`, `is_deobligation`,
`is_stray_uei`, …) and assigned `direction`.

**`$subawards_in`** — org × year inbound revenue
([dictionary](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/subawards_in.csv)): `subaward_in_amount`,
`n_subawards_in`. Folded into totals by `us_rollup()` as
`total_net = obligation_net + subaward_in − subaward_out`.

## 4. Code values: award types and families

The full award-type taxonomy, classified live from the package
(`us_classify_award_type()` over every code), with each type's footprint in
the 50-nonprofit pilot. A zero is informative: loan guarantees (08) and
insurance (09) exist in the taxonomy but never appear in this population.

```{r}
codes <- us_classify_award_type(us_award_type_codes())
d <- read.csv("data-dictionary/pilot-descriptives.csv")
n <- aggregate(cbind(actions = n_transactions, awards = n_awards) ~ award_type_code,
               data = d, sum)
tab <- merge(codes, n, by = "award_type_code", all.x = TRUE)
tab[is.na(tab)] <- 0
tab <- tab[order(tab$award_group, tab$award_family, tab$award_type_code),
           c("award_group", "award_family", "award_type_code", "award_type_label",
             "actions", "awards")]
names(tab) <- c("group", "family", "code", "label", "pilot actions", "pilot awards")
tab$`pilot actions` <- format(tab$`pilot actions`, big.mark = ",")
tab$`pilot awards`  <- format(tab$`pilot awards`,  big.mark = ",")
knitr::kable(tab, row.names = FALSE, align = c("l", "l", "l", "l", "r", "r"))
```

The families are the revenue-treatment switch:

| `award_family` | treatment |
|---|---|
| `grant` | revenue |
| `contract` | revenue |
| `other` | revenue-like (cooperative R&D) |
| `direct_payment` | usually excluded — student aid flowing *through* the institution |
| `loan` | obligation is \$0 by design; face value never touches revenue |
| `idv` | a vehicle: obligations only, never ceilings; money is on the delivery orders |

## 5. Code values: action types and classes

Action codes **collide across families** — the same letter means different
things — so classification always takes the family:

```{r}
knitr::kable(us_classify_action(c("A", "B", "C", "D"), "assistance"))
knitr::kable(us_classify_action(c("A", "B", "C", "D", "G", "K", "M"), "contract"))
```

`action_class` describes *intent*, never sign: a `continuation` can carry a
negative amount and a `termination` can carry zero. De-obligations are
identified from the amount's sign alone.

## 6. Code values: subaward direction

Direction exists only relative to an organization's UEI set, which is why
`us_normalize_subawards()` requires one:

| `direction` | meaning | panel treatment |
|---|---|---|
| `out` | the org is the prime — pass-through it owes onward | **subtracted**: `net_revenue = obligation_net − subaward_out` |
| `in` | the org is the subawardee — revenue invisible in prime data | added, in `subawards_in` (org × year) |
| `internal` | both sides belong to the org (e.g. NYU → NYU School of Medicine) | excluded from both flows |
| `unrelated` | neither side is the org (multi-recipient award now held elsewhere) | excluded |

## 7. The pilot, summarized at this grain

The 50-nonprofit pilot (FY2008–FY2025; 302,025 transactions; outbound
pass-through fetched per prime award), by award group, family, and type.
Static snapshot computed by
[`data-raw/make-pilot-descriptives.R`](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/data-raw/make-pilot-descriptives.R);
also available as [CSV](https://github.com/Nonprofit-Open-Data-Collective/usaspend/blob/main/vignettes/data-dictionary/pilot-descriptives.csv).

```{r}
d <- read.csv("data-dictionary/pilot-descriptives.csv")
## dollars in billions; "\\$" keeps pandoc from pairing the dollar signs
## across cells and reading the span between them as math
usd <- function(x) {
  s <- sprintf("\\$%.2f", abs(x) / 1e9)
  ifelse(x < 0 & round(abs(x) / 1e9, 2) > 0, paste0("-", s), s)
}
short <- c("Direct Payment for Specified Use"     = "Direct Payment (Specified)",
           "Direct Payment with Unrestricted Use" = "Direct Payment (Unrestricted)",
           "Other Financial Assistance"           = "Other Assistance",
           "IDC Indefinite Quantity"              = "IDC (Indef. Qty)",
           "Definitive Contract"                  = "Definitive")
type <- ifelse(d$award_type_label %in% names(short),
               short[d$award_type_label], d$award_type_label)
row <- function(fam, ty, x) data.frame(
  family = fam, type = ty,
  actions = format(x$n_transactions, big.mark = ","),
  awards  = format(x$n_awards, big.mark = ","),
  positive = usd(x$obligation_positive), negative = usd(x$obligation_negative),
  net = usd(x$obligation_net), passthrough = usd(x$subaward_out),
  `net revenue` = usd(x$net_revenue), check.names = FALSE)
tab <- rbind(
  row(d$award_family, type, d),
  row("", "**TOTAL**", as.data.frame(t(colSums(
    d[, c("n_transactions", "n_awards", "obligation_positive", "obligation_negative",
          "obligation_net", "subaward_out", "net_revenue")])))))
knitr::kable(tab, align = c("l", "l", rep("r", 7)))
```

Dollar figures are billions. The `family` column implies the award group:
`grant`, `direct_payment`, `loan`, and `other` are assistance awards;
`contract` and `idv` are contracts. (`awards` in the TOTAL row slightly
overcounts distinct awards, since a few span families.)

The table is the whole accounting argument in one place: project grants and
cooperative agreements carry both the dollars and nearly all the
pass-through; direct payments are student aid; loans net to zero obligations
by design; IDVs hold no money of their own. Total net revenue:
**\$`r format(round(sum(d$net_revenue) / 1e9, 1), nsmall = 1)`bn** on
\$`r format(round(sum(d$obligation_net) / 1e9, 1), nsmall = 1)`bn of net
obligations.

```{=html}
<script type="module">
  import mermaid from "https://cdn.jsdelivr.net/npm/mermaid@11/dist/mermaid.esm.min.mjs";
  mermaid.initialize({ startOnLoad: true, theme: "neutral" });
</script>
```
