---
title: "How USAspending data is structured"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How USAspending data is structured}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
ex <- us_sample_extract()
p  <- us_panel(ex)
```

USAspending publishes award data at three grains, and almost every analytic
mistake with this data comes from mixing them up.

| grain | one row is | carries |
|---|---|---|
| **transaction** | one award *action* — a new award, a modification, a correction | a signed dollar amount and a date |
| **award** | one prime award over its whole life | lifetime totals, current attributes, ceilings |
| **subaward** | one FSRS report line | prime → subrecipient flows (see `vignette("subawards")`) |

## An award is a ledger, not a payment

The award grain looks like the natural unit, but an award is a *container*.
Its life is written at the transaction grain — and most of that life is
paperwork:

```{r}
tx <- p$transactions
tx[award_key == "ASST_NON_HDTRA12310001_097",
   .(modification_number, action_date, action_type_label,
     federal_action_obligation)][order(action_date)]
```

Eight actions, four calendar years, three zero-dollar revisions. In the
sample as a whole, `r round(mean(tx$federal_action_obligation == 0, na.rm = TRUE) * 100)`%
of transactions carry $0 (in the contract-heavy sample; 23% in the 50-org
pilot) — administrative actions and period-of-performance changes. They are
the majority of *modifications* and they are how activity gets counted, so
the package keeps them, flagged.

**Only the transaction grain supports an annual panel.** The award grain has
one lifetime total and no dates money moved; the transaction grain has a
signed amount and a date on every action. That is the whole reason the
package's pipeline is built on transactions.

## The dozen dollar columns, in three groups

USAspending rows carry many dollar columns. Only one group is money received.

**Revenue — sum these.** `federal_action_obligation`, the signed amount
committed by each action, is the default measure (with
`pragmatic_obligation` as USAspending's loan-aware variant).

**Ceilings — never sum these.** `potential_total_value_of_award`,
`base_and_all_options_value`, `current_total_value_of_award`,
`total_funding_amount`. These describe *capacity*. The sample carries the
cautionary example — an indefinite-delivery vehicle with a $40,000,000
potential value and **zero dollars obligated** across its modifications:

```{r}
tx[award_key == "CONT_IDV_HHSD200200720032I_7523",
   .(action_date, award_type_label, federal_action_obligation)][order(action_date)][1:4]
sum(tx[award_key == "CONT_IDV_HHSD200200720032I_7523"]$federal_action_obligation)
```

A ceiling-based panel would credit this organization with $40M it never
received. (The money flows through the *delivery orders under* the vehicle,
which are separate awards — see `vignette("award-types")`.)

**Not revenue at all — separate columns.** `face_value_of_loan` (a
liability of the borrower), `original_loan_subsidy_cost` (the government's
expected loss), `non_federal_funding_amount` (someone else's cost share).

## Obligations are promises; outlays are cash; only promises are annual

`federal_action_obligation` is a *commitment*. The disbursement — the
outlay — happens later, often across years, and within any given year the
two differ **widely**. Measured on the FY2022+ VUMC awards whose
account-level (File C) records reconcile with their award transactions:
obligations put a median 90% of an award's dollars in its first fiscal
year while outlays put 2% there; the dollar-weighted center of cash lags
the center of commitment by roughly one year; and about two-thirds of an
award's dollars land in a *different year* under outlays than under
obligations. An obligations panel and a cash panel are different objects,
and neither approximates the other.

The award data carries **no annual outlay figure at all**:
`total_outlayed_amount_for_overall_award` is lifetime-to-date and blank on
most rows. Annual outlays exist only in the account-level File C data
(exposed per award at `POST /awards/funding/`, absent from every bulk
transaction file), and their coverage follows the reporting mandates:
quarterly and effectively optional from FY2017, required for
COVID-supplemental awards from April 2020, monthly and mandatory for all
agencies only from FY2022. Lifetime outlay totals therefore *undercount*
any award straddling those dates — pre-mandate disbursements were never
reported, not merely not yet paid.

Nor is coverage just an era problem. On the VUMC sample data, the share of
awards carrying a nonzero lifetime outlay:

| first action | assistance | contracts |
|---|---|---|
| pre-FY2017 | 33% | 0% |
| FY2017–21 | 78% | 15% |
| FY2022+ | 91% | 14% |

The contract gap is an *agency* gap: post-FY2022, 98% of HHS awards carry
an outlay against 0% of VA contracts and 2% of DoD awards. Whether outlays
exist for a portfolio depends on who funds it.

So an annual panel from this data measures obligations, and honest analysis
says so. The package enforces the honesty: `us_money_column("outlay")`
errors rather than substituting something close, and `us_reconcile()`
reports each award's obligation-to-outlay ratio so the size of the gap
stays visible. For portfolios where cash genuinely matters,
`us_add_outlays()` attaches annual File C outlays to a fiscal panel as a
*separate* column — fetched per award with `us_fetch_outlays()`, graded per
award for era, agency, and linkage coverage, and never substituted into
`obligation_net`.

## Award attributes are a moving target

Award-level fields — recipient, agency, even lifetime totals — are stamped
onto every transaction row *as of file generation*, and they change over an
award's life. Two consequences, both measured:

- **Take attributes from the latest action**, never from an arbitrary row —
  `us_normalize_awards()` does this. On many rows the award-level fields are
  simply blank.
- **The recipient itself can change.** NIH awards follow the principal
  investigator between institutions, so one award's history can span
  organizations. This is why the panel's grain is *organization* × award ×
  year, and why a UEI-filtered extract legitimately fails the lifetime
  identity on such awards (`vignette("reconciliation")`).

## Where to find every field

The reference companion to this vignette is `vignette("data-dictionary")`:
entity-relationship diagrams of the source database, the extract, and the
panel; keys and cardinality; the accounting-critical fields of every table;
the award-type, action-class, and direction code tables; and links to full
per-table dictionaries as CSV.

## Transactions come dirty

The raw ledger needs cleaning before it can be netted — duplicates from
overlapping download batches, correction and delete indicator codes,
county-level aggregate records with no identified recipient. The rules, each
grounded in measured pilot data, are the subject of
`vignette("accounting")`.
