---
title: "Detecting disruption in federal funding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting disruption in federal funding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7.2, fig.height = 4, dpi = 96)
options(datatable.print.trunc.cols = TRUE, width = 90)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

When federal funding is disrupted, the first question is usually "how much
money was taken back?" The ledger answers it badly. Terminations are
recorded as notices, the money often stays obligated on paper for months
or years, and much of what happens to a disrupted award is never formally
decided at all: the end date moves up, the next year's funding doesn't
arrive, the cash stops. This vignette is a tutorial on reading those
signals — which fields carry them, how to turn them into rates, and what
each kind of disruption looks like in a real award's history.

It is organized around a four-part taxonomy:

| type | what happens | where it shows up |
|---|---|---|
| **1. Termination** | the award ends early, formally or not | termination action codes and notice text; end dates cut to "now"; early closeouts |
| **2. Reduction** | the award continues, smaller | early de-obligations; ceiling cuts; shortened schedules; missing continuations |
| **3. Delay** | the award continues, later | extensions; administrative actions; late or missing funding actions; active awards with no outlays |
| **4. Transfer** | the work continues under another award, agency, or organization | re-awards; agency changes; transfer actions; novations |

Everything below runs offline on `disruption_sample`, a bundled set of
aggregates and case ledgers from a 1,000-nonprofit pull.

```{r setup, message = FALSE}
library(usaspend)
library(data.table)
d <- disruption_sample
str(d$meta[c("sample", "data_end", "t0", "n_orgs_with_awards", "n_awards",
             "n_transactions", "largest_recipient_share")])
```

**Read the numbers as a worked example, not an estimate.** The sample is
the first 1,000 rows of a nonprofit crosswalk ordered by state, and one
research contractor accounts for
`r round(100 * d$meta$largest_recipient_share)`% of its transactions.
The method generalizes; the magnitudes belong to this portfolio.

## The finding that organizes everything else

Between February and August 2025, formal contract terminations in this
sample went from roughly zero a year to
`r d$signals[year == 2025, term_code]`. De-obligated dollars did not follow:

```{r headline}
knitr::kable(d$signals[year >= 2021, .(year,
  term_code_actions = term_code, term_notice_actions = term_text,
  deobligated_M = round(deob_dollars / 1e6, 1),
  early_deob_M  = round(early_deob_dollars / 1e6, 1),
  end_date_cuts = end_cut, ceiling_cut_M = round(ceiling_cut_dollars / 1e6, 1))])
```

(The 2024 de-obligation figure is one COVID-19 trial closing out; it is
not a policy signal.) The reason is visible in the terminated awards
themselves:

```{r headline-money}
t <- d$terminations
t[kind == "formal", .(
  awards = .N,
  notice_was_zero_dollars = sum(change_at_notice == 0),
  still_obligated = sum(status == "still obligated"),
  median_days_to_deobligation = median(as.numeric(first_deob_after - t_date), na.rm = TRUE))]
```

Almost every termination notice moved **no money**. Most terminated
awards were still carrying their obligations at the end of the data; where
money was eventually pulled back, it took a median of more than six months.
A detector built on de-obligations misses most terminations entirely, and
dates the rest by the paperwork rather than the decision.

The practical consequence: **measure disruption from the non-monetary
fields** — action codes, end dates, ceilings, and description text — and
compare their *rates* to a pre-period baseline.

## Getting the fields

Four fields that carry these signals are part of the canonical schema:
`transaction_description` (the free text on each action),
`base_and_all_options_value` (the change to a contract's ceiling on this
action), `pop_potential_end_date`, and the awarding office
(`awarding_office_code`, `awarding_office_name`). `us_disruption_flags()`
reads them and adds one logical column per signal:

```{r build, eval = FALSE}
ex <- us_extract(my_ueis, years = 2016:2026)
tx <- us_normalize_transactions(ex$transactions, requested_uei = my_ueis)
f  <- us_disruption_flags(tx)
names(f)[grepl("^dsr_", names(f))]
```

```{r flag-names, echo = FALSE}
grep("^dsr_", names(d$cases), value = TRUE)
```

Each flag is a literal property of one action, so every count can be
audited against the description text. None of them means "disruption" on
its own — closeouts happen every year, and end dates move constantly on
some kinds of award. Disruption is a **change in the rate**.

Three rules make the rates comparable:

- **Compare the same months.** Obligations spike every September; every
  table here compares February–August of each year, the window after the
  inauguration that all years share.
- **Remove routine churn.** Supply delivery orders (Defense Logistics
  Agency purchase orders) cancel and re-date constantly; the bundled
  tables exclude them.
- **Mind reporting lag.** Contract actions can post up to ~90 days late,
  assistance up to ~30. The most recent months are always incomplete.

## The ten-year baseline

A disruption is measured against a trend, and the trend needs to be long
enough to contain the last shock: COVID-19 relief ran through FY2020–FY2021.
Gross obligations (new money committed, before claw-backs) by the
agency that originated each award, over the same October–August window of
each fiscal year:

```{r ggplot, eval = has_ggplot, include = FALSE}
suppressWarnings(library(ggplot2))
```

```{r trend-plot, eval = has_ggplot, fig.height = 5.2}
tr <- d$trend_fy[agency %in% c("HHS", "USAID", "DoD", "Education", "EPA", "USDA")]
tr[, agency := factor(agency, c("HHS", "USAID", "DoD", "Education", "EPA", "USDA"))]
ggplot(tr, aes(fy, gross_obligations / 1e6)) +
  annotate("rect", xmin = 2019.5, xmax = 2021.5, ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_vline(xintercept = 2024.5, colour = "grey55", linetype = "dashed") +
  geom_line(colour = "#1a5276", linewidth = 0.7) +
  geom_point(colour = "#1a5276", size = 1.4) +
  facet_wrap(~ agency, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = c(2016, 2020, 2024)) +
  labs(title = "Gross obligations by originating agency, FY2016-FY2026",
       subtitle = "October-August of each fiscal year; shaded: COVID-19 relief years; dashed: FY2025 begins",
       x = NULL, y = "$ millions (free scale per panel)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

```{r trend-table}
tot <- d$trend_fy[, .(gross_M = round(sum(gross_obligations) / 1e6)), by = fy]
tot[, change_vs_FY2024 := sprintf("%+.0f%%", 100 * (gross_M / gross_M[fy == 2024] - 1))]
tot
```

Three things the trend makes visible that a two-year comparison would not.
The COVID-19 years are a bump, not a new level, so a drop measured against
FY2021 alone would be understated. FY2023–FY2024 were this portfolio's
peak: FY2025's fall only returns it to about its COVID-era level, while
FY2026 is below FY2019. And FY2025 straddles the transition — its
October–January months ran at the old pace (the EPA spike is almost
entirely one $95M agreement signed on 8 January 2025, which reappears below
as a quiet truncation), so the full effect lands in FY2026.

Monthly, with the inauguration marked:

```{r monthly-plot, eval = has_ggplot, fig.height = 3.4}
m <- d$trend_month[month >= as.Date("2021-10-01")]
m[, roll12 := frollmean(gross_obligations, 12)]
ggplot(m, aes(month)) +
  geom_col(aes(y = gross_obligations / 1e6), fill = "#c7d6e2", width = 25) +
  geom_line(aes(y = roll12 / 1e6), colour = "#1a5276", linewidth = 0.8, na.rm = TRUE) +
  geom_vline(xintercept = d$meta$t0, colour = "grey40", linetype = "dashed") +
  labs(title = "Gross obligations by month, with 12-month rolling mean",
       subtitle = "dashed: 2025-01-20. Latest months are incomplete (reporting lag)",
       x = NULL, y = "$ millions") +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
```

## 1. Termination

A termination ends an award before its scheduled end. It appears in three
forms, from most to least formal.

**Formal termination.** Contracts have action codes for it: `F`
(terminate for convenience), `E` (for default), `N` (legal cancellation)
— `dsr_term_code`. Grants have no termination code; the only formal
record is the notice text in `transaction_description` — `dsr_term_text`.
Test for both, and exclude "determination".

**Quiet truncation.** The end date moves to (about) today with no notice
and, usually, no money — `dsr_end_cut`. In this sample these outnumber
formal terminations on assistance awards.

**Early wind-down.** A spike in contract closeouts (`K`, `dsr_closeout`)
and in assistance "adjustment to completed award" actions: awards being
finished off ahead of schedule.

```{r term-signals}
d$signals[year >= 2021, .(year, term_code, term_text, end_cut,
                          end_cut_with_no_money = end_cut_no_deob,
                          closeout, rescinded)]
```

Timing separates the two forms sharply. Formal notices came in a wave in
the first months and then stopped; quiet truncations continued at a steady
rate through 2026:

```{r term-month, eval = has_ggplot, fig.height = 3.4}
tm <- t[, .N, by = .(month = as.Date(format(t_date, "%Y-%m-01")), kind)]
ggplot(tm, aes(month, N, fill = kind)) +
  geom_col(position = position_dodge2(preserve = "single", padding = 0.15), width = 25) +
  scale_fill_manual(values = c(formal = "#1a5276", quiet_truncation = "#9a6a1f"),
                    labels = c("formal (code or notice)", "quiet truncation (end date cut)"),
                    name = NULL) +
  labs(title = "Terminations by month of first signal",
       x = NULL, y = "awards") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

What happened to the money afterwards:

```{r term-status}
dcast(t, kind ~ status, fun.aggregate = length)
t[, .(awards = .N, obligated_before_M = round(sum(obligated_before) / 1e6, 1),
      pulled_back_M = round(-sum(deobligated_since + pmin(change_at_notice, 0)) / 1e6, 1),
      ceiling_cut_M = round(-sum(ceiling_cut) / 1e6, 1)), by = kind]
```

**Recipe — count terminations.** Take the first formal signal per award
after the reference date; add awards whose schedule was cut without one.

```{r term-recipe, eval = FALSE}
T0 <- as.Date("2025-01-20")
formal <- f[action_date >= T0 & (dsr_term_code | dsr_term_text),
            .(t_date = min(action_date), kind = "formal"), by = award_key]
quiet  <- f[action_date >= T0 & dsr_end_cut & !award_key %in% formal$award_key,
            .(t_date = min(action_date), kind = "quiet_truncation"), by = award_key]
events <- rbind(formal, quiet)
```

### Case: a notice and nothing else

A USAID energy contract. Four years of incremental funding, then a single
zero-dollar action — and no further record. At the end of the data the
award's full obligation still stands.

```{r case-helper}
show_case <- function(k, from = "1900-01-01") {
  x <- d$cases[case == k & action_date >= as.Date(from), .(action_date, mod = modification_number,
                            type = action_type_label,
                            obligation = round(federal_action_obligation),
                            ceiling = round(ceiling_change), end = pop_end_date,
                            description = substr(transaction_description, 1, 90))]
  knitr::kable(x, format.args = list(big.mark = ","))
}
show_case("termination_no_deob")
```

### Case: notice first, money months later

The 2019–20 National Postsecondary Student Aid Study. The notice on
10 February 2025 cut the end date from 2028 to that day and moved no money;
the termination settlement five months later de-obligated $5.9M and cut
the ceiling by $33.6M. A detector keyed on de-obligations would date this
termination to July.

```{r case-deob}
show_case("termination_deob", from = "2024-01-01")
```

### Case: terminated, reinstated — smaller

Its successor study, NPSAS:24, was terminated the same day and de-obligated
within a week ($10.2M). In June the award was **reinstated** — with new
money, but its ceiling cut by $51M and its schedule shortened from 2032 to
2027. Rescissions (`dsr_rescinded`) are rare but must be tracked: the
termination did not stick, the reduction did.

```{r case-rescinded}
show_case("termination_rescinded", from = "2024-12-01")
```

## 2. Reduction

A reduction leaves the award alive and smaller. The formal instrument is
the de-obligation, but most reductions are recorded some other way.

- **Early de-obligation** (`dsr_early_deob`): money withdrawn while the
  award still has more than 60 days to run. Distinguish it from the routine
  closeout de-obligation *after* the end date, which every year produces.
- **Ceiling cut** (`dsr_ceiling_cut`): a contract's potential value
  (base plus all options) reduced. Future work removed, often with no money
  moving at all.
- **Shortened schedule** (`dsr_end_cut`): the same flag as a quiet
  termination when the new end is "now"; a reduction when it is later.
- **Fewer continuations and options**: multi-year grants are funded a year
  at a time (continuation actions), contracts by exercising options. Fewer
  of either is a reduction that leaves no row at all — it is visible only
  as a drop in the rate.

```{r reduction}
d$signals[year >= 2021, .(year, early_deob, early_deob_M = round(early_deob_dollars / 1e6, 1),
                          ceiling_cut, ceiling_cut_M = round(ceiling_cut_dollars / 1e6, 1),
                          end_cut)]
d$action_types[action_type_label %in% c("Continuation", "New", "Exercise an Option",
                                        "Funding Only Action")]
```

Early de-obligations barely moved; ceiling-cut dollars were about four
times their 2021–2024 average, and continuation actions fell by a third. The cuts were made to *future*
money.

### Case: a ceiling cut to zero

An NIH communications support agreement (a blanket purchase agreement,
under which orders are placed). In July 2026 its entire $49.9M ceiling was
removed in a zero-dollar administrative action — nothing had been obligated
under it, so no de-obligation was possible. The description cites an
executive-order code.

```{r case-ceiling}
show_case("ceiling_cut")
```

### Case: a quiet truncation with $95M left in place

An EPA assistance agreement awarded on 8 January 2025 for $95M through
2027. On 30 April 2025 a zero-dollar action set its end date to that day.
No termination language, no de-obligation: in the obligation ledger this
award still reads as $95M committed.

```{r case-quiet}
show_case("quiet_truncation")
```

## 3. Delay

Delays are the least formal category and the most informative about
behavior that departs from formal decisions: awards that are not
terminated or reduced, but whose money arrives late or not at all.

**Administrative actions and extensions.** Zero-dollar administrative
actions (`dsr_admin`) are where stop-work orders and similar notices are
usually filed. Extensions (`dsr_extension`), which one might expect to rise,
*fell* — agencies were not extending the awards they were slowing down.

```{r delay-signals}
d$signals[year >= 2021, .(year, admin, zero_dollar, extension, stop_work)]
```

**Late or missing funding actions.** A multi-year grant receives its next
year's money roughly a year after the last. Anchor on each funded action
of an award with more than 400 days left to run, and ask whether the next
funding action arrived on time:

```{r continuations}
cn <- d$continuations[, .(N = sum(N)), by = .(due_year, outcome)]
cn <- dcast(cn, due_year ~ outcome, value.var = "N", fill = 0)
cn[, share_not_on_time := round(1 - `on time` / (`on time` + `30-120 days late` +
                                                 `missing or >120 days late`), 3)]
cn
dcast(d$continuations[agency %in% c("HHS", "USAID", "Education"),
                      .(late = round(sum(N[outcome != "on time"]) / sum(N), 2)),
                      by = .(agency, due_year)], agency ~ due_year, value.var = "late")
```

Overall the not-on-time share barely moved, but two things changed under
it: the USAID series jumps, and the moderately late category (30–120 days)
more than doubled in 2025. The aggregate hides both because most anchors
belong to programs that kept paying on schedule.

**Active awards with no cash.** The most direct evidence of a payment
delay is an award that is scheduled to be running, has not been terminated
on paper, and receives no outlays. File C account-level data
(`us_fetch_outlays()`) reports outlays by period for awards starting FY2020
onward. For each quarter, take the awards whose schedule — as known at the
start of the quarter — covers the whole quarter, and count those with no
outlay:

```{r outlays}
oq <- d$outlay_quarters[fy * 10 + fq < d$meta$outlay_last_quarter[["fy"]] * 10 +
                                          d$meta$outlay_last_quarter[["fq"]]]
all_q <- oq[, .(active = sum(active), no_cash_share = round(sum(no_cash) / sum(active), 3)),
            by = .(fy, fq)][order(fy, fq)]
all_q
```

```{r outlays-plot, eval = has_ggplot, fig.height = 3.8}
oa <- oq[agency %in% c("HHS", "USAID", "EPA", "Education"),
         .(share = sum(no_cash) / sum(active), active = sum(active)), by = .(agency, fy, fq)]
oa[, q := fy + (fq - 1) / 4]
ggplot(oa, aes(q, share)) +
  geom_vline(xintercept = 2025.25, colour = "grey55", linetype = "dashed") +
  geom_line(colour = "#1a5276", linewidth = 0.7) +
  geom_point(colour = "#1a5276", size = 1.3) +
  facet_wrap(~ agency, ncol = 4) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%"), limits = c(0, 1)) +
  scale_x_continuous(breaks = 2023:2026) +
  labs(title = "Share of scheduled-active awards receiving no outlay in the quarter",
       subtitle = "File C outlays, awards first obligated FY2020+; dashed: FY2025 Q2 (Jan-Mar 2025)",
       x = "fiscal year", y = NULL) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
```

Two different disruptions are visible here, and the method separates them.
USAID's no-cash share rose to nearly 100% and stayed there: awards that
were still open on paper stopped being paid. The all-agency spike in FY2026
Q1 (October–December 2025) coincides with the fall 2025 lapse in
appropriations and reverses the next quarter: a system-wide timing
disruption, not a program decision. The share also moves with the season,
which is why every comparison should be same-quarter.

### Case: the continuation that became two revisions

An NIH early-career grant, funded each January. The January 2025
continuation never came; instead, two zero-dollar revisions in May and June
moved the end date out a year. The award was neither terminated nor
reduced — its money was delayed, and the schedule absorbed it.

```{r case-continuation}
show_case("missed_continuation")
```

## 4. Transfer

A transfer moves the work rather than ending it. Whether the ledger can
link the old and new award depends on who the work moves to.

| the work moves to | same award id? | how to find it |
|---|---|---|
| another agency (agency closure) | **yes** | `dsr_agency_change`; `dsr_transfer` action pairs |
| another organization, contract | yes | novation, `dsr_novation` (action `J`) |
| another organization, NIH grant | yes (FAIN persists) | recipient change on the same award key |
| a new award, same organization | no | match on recipient + office + program + text |
| a new award, another organization | no | match on office + program + text, across all recipients |

### Case: re-awarded to the same organization

The National Youth Tobacco Surveys contract was terminated on 1 May 2025
— the notice text cites the DOGE cost-efficiency initiative. Four months
later the same organization received a new award for the same survey,
from a different HHS sub-agency, at about a third of the old ceiling.

```{r case-reaward}
show_case("reaward_old", from = "2024-12-01")
show_case("reaward_new")
```

Nothing in either record points to the other; the new award does not cite
the old number. Linking them is a matching problem. Among same-recipient
awards made from 90 days before to a year after each termination, the
bundled table scores candidates on description similarity, same office,
same program code, and whether the new award's text cites the old award id:

```{r reaward-links}
t[, .N, by = link][order(-N)]
knitr::kable(t[link == "probable", .(t_date, recipient = substr(recipient, 1, 26),
                                     old_award = award_key, successor, successor_start)])
```

Shared office and program code alone are **not** evidence — contract
vehicles issue many orders from one office — so a "probable" link requires
closely matching description text or an explicit citation. Few terminations
in this sample found a same-organization successor.

### Case: the NIH grant that followed its investigator

When a principal investigator moves, an NIH grant moves with them: same
award number, new recipient. A pull filtered on one organization's UEI sees
only its own share of the history. These grants appear as reconciliation
breaks in the original holder's panel (`vignette("reconciliation")`):

```{r pi-transfers}
knitr::kable(d$pi_transfers[, .(award_key, reported_total, in_sample_total,
                                other_recipients = substr(other_recipients, 1, 70))],
             format.args = list(big.mark = ","))
```

### Case: novation

A novation keeps the contract and replaces the contractor — for example
after an acquisition. It is one action with its own code:

```{r case-novation}
show_case("novation")
```

Awards re-made to a *different* organization are invisible from inside a
UEI-filtered sample: finding them means querying all recipients for new
awards from the same office and program after each termination.

## When an agency closes: what persists and what changes

USAID's contracts and grants did not disappear from USAspending when the
agency was wound down in 2025. The sample holds
`r d$usaid$awards` USAID awards, `r d$usaid$active_at_T0` of them still
scheduled to run on 20 January 2025:

```{r usaid}
d$usaid[c("awards", "active_at_T0", "with_action_since_T0", "terminated",
          "terminated_deobligated", "moved_to_state")]
knitr::kable(d$usaid$actions_since_T0)
```

Two records show the mechanics. An energy contract vehicle (an IDIQ) and one
of its task orders, both held by the same organization:

```{r case-usaid}
show_case("usaid_idv", from = "2024-11-01")
knitr::kable(d$cases[case == "usaid_task_order" & action_date >= "2025-10-01",
  .(action_date, mod = modification_number, type = action_type_label,
    obligation = round(federal_action_obligation), awarding_agency_name,
    funding_agency_name, description = substr(transaction_description, 1, 70))],
  format.args = list(big.mark = ","))
```

The vehicle received a termination notice in February 2025, had it
rescinded in December, and appeared under the Department of State in
February 2026. The task order was extended, given $12.3M of new work, and
in August 2026 moved to State with a pair of equal and opposite actions: a
**Transfer Action** of −$14.1M and an administrative action of +$14.1M.

What this shows about the database logic:

**Persists:**

- **The award key.** USAspending's generated unique award id
  (`award_key`) is built when the award is created, from the award number
  and the awarding agency's code — `CONT_AWD_{PIID}_{agency}_{parent
  PIID}_{parent agency}` for contracts, `ASST_NON_{FAIN}_{agency}` for
  assistance. The `7200` in these contract keys is USAID's code, and it stays
  there after the move. The key is an identifier, not a statement of who manages
  the award now.
- **The award number** (PIID/FAIN), the parent-vehicle link, the
  modification sequence (P00013, P00014 continue the same series), and the
  recipient.
- **The funding agency, where money moved.** On the transferred task order
  the *funding* agency is still USAID: the money was appropriated to
  USAID's accounts. (On the vehicle itself, which holds no money, both the
  awarding and funding fields changed to State.)

**Changes:**

- **The awarding agency, sub-agency, and office** on each action *after*
  the move. Earlier actions keep USAID. An award's agency is therefore a
  property of each action, not of the award.
- **Vehicle and orders move separately.** The IDV and its task order moved
  six months apart.
- **The money moves by transfer action.** The −/+ pair nets to zero but
  inflates gross de-obligation counts; remove transfer pairs (or net by
  award and date) before measuring reductions.

**Implications for analysis:**

- Decide whether an award belongs to the agency that **originated** it
  (first action, or the code in the key) or the one that **manages** it
  now (latest action). The bundled tables use the originating agency, so
  USAID's collapse shows as USAID's; a latest-action rollup would show a
  smaller USAID and a slightly larger State.
- New work that State re-competes will be **new awards with new keys**,
  under State's agency code — with no link to the USAID award it replaces.
  Tracing that is the cross-organization matching problem above.
- Grouping by the key's agency code, by `awarding_agency_name`, and by
  `funding_agency_name` gives three different "USAID portfolios". Say which
  one you mean.

## Transaction types worth tracking

Counts of each action type over February–August, with the 2025 rate
relative to the 2023–2024 average:

```{r action-types}
at <- copy(d$action_types)
at[, ratio_2025 := round(`2025` / pmax((`2023` + `2024`) / 2, 1), 2)]
knitr::kable(at[order(award_group, -ratio_2025)])
```

In rough order of usefulness:

| signal | field / flag | why it matters |
|---|---|---|
| terminate for convenience / default / cancellation | `action_type_code` F, E, N; `dsr_term_code` | the formal record of a termination, for contracts |
| termination notice text | `transaction_description`; `dsr_term_text` | the *only* formal record for grants; also catches notices filed under other codes |
| end date cut | `pop_end_date` vs previous; `dsr_end_cut` | quiet termination and reduction — often with no money and no notice |
| ceiling cut | `base_and_all_options_value` < 0; `dsr_ceiling_cut` | future work removed; invisible in obligations |
| missing continuations and option exercises | assistance `B`, contract `G` rates | reductions that leave no row; measure as a rate |
| no outlays on active awards | File C, `us_fetch_outlays()` | payment delay behind unchanged paperwork |
| closeouts and adjustments to completed awards | contract `K`, assistance `D`; `dsr_closeout` | early wind-down |
| rescissions | `dsr_rescinded` | terminations that did not stick |
| administrative actions | contract `M`; `dsr_admin`, `dsr_stop_work` | stop-work orders and notices |
| transfer, novation, agency change | contract `T`, `J`; `dsr_agency_change` | work moved, not ended |
| early de-obligation | `dsr_early_deob` | the formal reduction — slow, and the lagging indicator |

## Limits

- **A UEI-filtered sample cannot see the other side of a transfer**:
  re-awards to another organization, and the share of a moved grant held
  elsewhere, require queries by office and program across all recipients.
- **Recent months are incomplete.** Reporting lag understates the latest
  quarter; the File C quarter still being reported is excluded above.
- **Text signals depend on agency practice.** Some agencies write "notice
  of termination"; some write nothing. Always pair text with codes and
  schedule changes.
- **File C covers awards from FY2020** and only agencies that report
  outlays (DoD reports almost none), so the no-cash measure is a
  subsample.
- **Flags are evidence, not conclusions.** Audit every count against the
  description text before reporting it.
