---
title: "Fitting an outlay-imputation model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting an outlay-imputation model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7.2, fig.height = 4, dpi = 96)
options(datatable.print.trunc.cols = TRUE, width = 90)
```

The bundled `outlay_model` was fitted on 1,184 ground-truth awards from 51
nonprofits. If your portfolio looks different — other agencies, other award
durations, another era — refit on your own ground truth. The whole pipeline
is three calls: **retrieve** (`us_outlay_training()`), **inspect/extend
features** (`us_outlay_features()`), **fit and evaluate**
(`us_impute_fit()`, `us_impute_eval()`).

```{r setup, message = FALSE}
library(usaspend)
library(data.table)
```

## 1. Retrieve: build a training set

`us_outlay_training()` does the whole ground-truth screen: award features
from your transaction ledger, annual File C outlays per award (fetched via
`us_fetch_outlays()`, one paged request per candidate — sustained pulls
beyond ~1,000 requests can trip a temporary host block, so large pulls are
worth caching), linkage, and the two truth tiers.

```{r fetch, eval = FALSE}
tx <- us_normalize_transactions(my_extract$transactions)
tr <- us_outlay_training(tx)          # fetches File C for every candidate
# or, with a cached pull:
tr <- us_outlay_training(tx, funding = my_cached_funding)
```

This vignette works offline with the bundled set:

```{r training}
tr <- outlay_training
tr
tr$awards[!is.na(tier), .N, by = .(tier, mod_class)][order(tier, -N)]
```

The screens are deliberate, not incidental — an award whose File C records
don't reconcile with its own ledger, or whose cash story is censored,
teaches the model the reporting system instead of the spending. Tiers,
thresholds and rationale are specified in `IMPUTATION.md` §1.

## 2. Features

`us_outlay_features()` is the shared feature layer — the training builder
calls it, and `us_impute_outlays()` calls it again at prediction time, so
anything you compute here is available on both sides.

```{r features, message = FALSE}
f <- us_outlay_features(us_normalize_transactions(vumc_transactions))
f[, .N, by = .(dur_bin, late_start)][order(dur_bin, late_start)][1:8]
```

The default model cells are `dur_bin` (award duration in fiscal years,
capped at 6) × `late_start` (first obligated April–September). The
experiment tested award *type* as a third dimension and found it added
nearly nothing beyond duration in a grant-dominated sample — but the
`cells` argument takes any feature columns present in the grid, so a
portfolio where type matters can use them.

## 3. Fit

```{r fit}
m <- us_impute_fit(tr, cells = c("dur_bin", "late_start"), min_cell = 8)
m
```

The fitted object is transparent — plain curve tables you can inspect and
plot. The duration-4 curves show both regularities the model banks on: the
lag (mass at years 1–2) and the late-start shift (the `TRUE` curve barely
spends in year 0):

```{r curves-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
cc <- m$curves_cell[dur_bin == 4 & n >= m$min_cell]
ggplot(cc, aes(t, share, linetype = late_start)) +
  geom_line(colour = "#1a5276", linewidth = 0.8) +
  geom_point(colour = "#1a5276", size = 2) +
  labs(title = "Duration-4 liquidation curves",
       subtitle = "share of net obligations outlaid per event-year; the curve sums to the cell's outlay/obligation ratio",
       x = "years since first obligation", y = "share of net obligations",
       linetype = "late start (Apr–Sep)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## 4. Evaluate

`us_impute_eval()` refits out-of-fold and scores against the two reference
rules on both metrics — `timing` (misallocation share, level normalized
away) and `level_timing` (the method is charged for missing the cash level
too).

```{r eval, message = FALSE}
ev <- us_impute_eval(tr, folds = 5)
ev$summary
ev$by_class
```

Read `ev$by_class` before trusting the model on a skewed portfolio:
`reduced` awards are the hardest class for every method, and a portfolio
dominated by them deserves wider error bars, not a better point estimate.

## 5. Use it

A fitted model drops into the imputation functions anywhere the bundled
one would be used:

```{r use, message = FALSE}
tx  <- us_normalize_transactions(us_sample_extract()$transactions)
imp <- us_impute_outlays(tx, model = m)
imp[, .(dollars = round(sum(outlay_imputed))), by = imputation_method]
```

```{r save, eval = FALSE}
saveRDS(m, "outlay-model-2026.rds")          # reuse across sessions
p <- us_panel(my_extract, period = "fiscal")
p <- us_add_imputed_outlays(p, model = m)
```

Refit when your ground truth grows — every fiscal year that closes moves
more awards through the truth screens, and the support envelope (the
durations the model will speak for) widens with it.
