---
title: "Fitting an outlay-imputation model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting an outlay-imputation model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7.2, fig.height = 4, dpi = 96)
options(datatable.print.trunc.cols = TRUE, width = 90)
```

The bundled `outlay_model` was fitted on 1,184 ground-truth awards from 51
nonprofits. If your portfolio looks different — other agencies, other award
durations, another era — refit on your own ground truth. The whole pipeline
is three calls: **retrieve** (`us_outlay_training()`), **inspect/extend
features** (`us_outlay_features()`), **fit and evaluate**
(`us_impute_fit()`, `us_impute_eval()`).

```{r setup, message = FALSE}
library(usaspend)
library(data.table)
```

## 1. Retrieve: build a training set

`us_outlay_training()` does the whole ground-truth screen: award features
from your transaction ledger, annual File C outlays per award (fetched via
`us_fetch_outlays()`, one paged request per candidate — sustained pulls
beyond ~1,000 requests can trip a temporary host block, so large pulls are
worth caching), linkage, and the two truth tiers.

```{r fetch, eval = FALSE}
tx <- us_normalize_transactions(my_extract$transactions)
tr <- us_outlay_training(tx)          # fetches File C for every candidate
# or, with a cached pull:
tr <- us_outlay_training(tx, funding = my_cached_funding)
```

This vignette works offline with the bundled set:

```{r training}
tr <- outlay_training
tr
tr$awards[!is.na(tier), .N, by = .(tier, mod_class)][order(tier, -N)]
```

### How the retrieval filters, and why

The screens are deliberate, not incidental — an award whose File C records
don't reconcile with its own ledger, or whose cash story is censored,
teaches the model the reporting system instead of the spending. In order:

1. **Candidates**: first obligated FY2020+ with material net obligations.
   The floor is not arbitrary — account-level outlay reporting became
   mandatory (monthly) only in **FY2022**, with COVID awards from April
   2020 and effectively optional coverage before. Earlier awards have
   permanently unreported cash and *cannot* be ground truth.
2. **Linkage**: File C lifetime obligations must reconcile with the
   award's own transaction ledger within 10%. Unlinkable File C records
   describe money that cannot be tied to the award.
3. **A finished cash story**: either lifetime outlays ≈ net obligations
   with cash no longer flowing (`tier = "reconciled"`), or a
   mandate-era start with a completed, plateaued series
   (`tier = "shape_complete"` — the level may fall short but the timing
   profile is fully observed).

The sample frame is therefore **shaped by the 2022 mandate**: recent,
shorter awards from agencies that report. That has two consequences worth
stating before fitting. First, the *support envelope* — long awards are
scarce (few 5+ year awards have completed inside the window), so the
model declines to speak for them. Second, **agency selection**: outlay
coverage varies enormously by agency — DoD reports File C obligations but
essentially no outlays (median outlay/obligation ratio 0.00), Commerce
0.24, versus NSF, USDA, and NASA at 1.00 and HHS at 0.91. Agencies that
don't report cash contribute *no* ground truth, so the fitted curves are
estimated from the reporters. If the level of outlay reporting is
correlated with your outcome of interest, and liquidation shapes differ
across agencies, the imputed values carry that selection signature — a
DoD-heavy portfolio is being imputed with curves learned mostly from
HHS-and-NSF-style awards. Check your portfolio's agency mix against the
training mix (`tr$awards[!is.na(tier), .N, by = awarding_agency_name]`)
before leaning on the imputations, and consider agency cells
(`cells = c("dur_bin", "late_start", ...)`) where your ground truth is
deep enough to support them.

Tiers, thresholds and rationale are specified in `IMPUTATION.md` §1.

## 2. Features

`us_outlay_features()` is the shared feature layer — the training builder
calls it, and `us_impute_outlays()` calls it again at prediction time, so
anything you compute here is available on both sides.

```{r features, message = FALSE}
f <- us_outlay_features(us_normalize_transactions(vumc_transactions))
f[, .N, by = .(dur_bin, late_start)][order(dur_bin, late_start)][1:8]
```

The default model cells are `dur_bin` (award duration in fiscal years,
capped at 6) × `late_start` (first obligated April–September). The
experiment tested award *type* as a third dimension and found it added
nearly nothing beyond duration in a grant-dominated sample — but the
`cells` argument takes any feature columns present in the grid, so a
portfolio where type matters can use them.

## 3. Fit

```{r fit}
m <- us_impute_fit(tr, cells = c("dur_bin", "late_start"), min_cell = 8)
m
```

The fitted object is transparent — plain curve tables you can inspect and
plot. The duration-4 curves show both regularities the model banks on: the
lag (mass at years 1–2) and the late-start shift (the `TRUE` curve barely
spends in year 0):

```{r curves-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
cc <- m$curves_cell[dur_bin == 4 & n >= m$min_cell]
ggplot(cc, aes(t, share, linetype = late_start)) +
  geom_line(colour = "#1a5276", linewidth = 0.8) +
  geom_point(colour = "#1a5276", size = 2) +
  labs(title = "Duration-4 liquidation curves",
       subtitle = "share of net obligations outlaid per event-year; the curve sums to the cell's outlay/obligation ratio",
       x = "years since first obligation", y = "share of net obligations",
       linetype = "late start (Apr–Sep)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## 4. Evaluate

`us_impute_eval()` refits out-of-fold and scores against the two reference
rules on both metrics — `timing` (misallocation share, level normalized
away) and `level_timing` (the method is charged for missing the cash level
too).

```{r eval, message = FALSE}
ev <- us_impute_eval(tr, folds = 5)
ev$summary
ev$by_class
```

Read `ev$by_class` before trusting the model on a skewed portfolio:
`reduced` awards are the hardest class for every method, and a portfolio
dominated by them deserves wider error bars, not a better point estimate.

## 5. Use it

A fitted model drops into the imputation functions anywhere the bundled
one would be used:

```{r use, message = FALSE}
tx  <- us_normalize_transactions(us_sample_extract()$transactions)
imp <- us_impute_outlays(tx, model = m)
imp[, .(dollars = round(sum(outlay_imputed))), by = imputation_method]
```

```{r save, eval = FALSE}
saveRDS(m, "outlay-model-2026.rds")          # reuse across sessions
p <- us_panel(my_extract, period = "fiscal")
p <- us_add_imputed_outlays(p, model = m)
```

Refit when your ground truth grows — every fiscal year that closes moves
more awards through the truth screens, and the support envelope (the
durations the model will speak for) widens with it.
