---
title: "Obligations versus outlays: the two ledgers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Obligations versus outlays: the two ledgers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(datatable.print.trunc.cols = TRUE, width = 90)
```

```{r setup, message = FALSE}
library(usaspend)
library(data.table)
```

Federal spending is recorded twice, in two ledgers that answer different
questions:

* The **budget ledger** records **obligations** — binding commitments
  against budget authority, booked the day the award action is signed.
  This is what USAspending's award and transaction tables carry, what
  `us_panel()` nets, and what `us_reconcile()` can verify to the dollar.
* The **payment ledger** records **outlays** — Treasury disbursements
  that liquidate those commitments, arriving on their own schedule, often
  years later.

In principle the two converge: money obligated is eventually either paid
or de-obligated, so lifetime outlays equal lifetime net obligations at
closeout. Measured on this package's ground truth, awards with full
reporting coverage and two-plus years to liquidate hit a median
outlay/obligation ratio of exactly 1.00. Within any given *year*, though,
the ledgers diverge widely — obligations put a median 90% of an award's
dollars in its first fiscal year, cash puts 2% there — and the published
payment ledger has serious measurement problems that the budget ledger
does not.

## What changed in 2022

Account-level reporting of award spending (the "File C" submissions
agencies make under the DATA Act) followed a staged mandate:

* **FY2017–FY2021** — quarterly submissions, with outlay reporting
  effectively optional; most agencies reported obligations but not cash.
* **April 2020** — awards funded by a COVID-19 supplemental were required
  to report outlays monthly.
* **FY2022** — all agencies required to submit outlay data for award
  spending **every month**.

The consequence is a hard information boundary: outlay history before an
award's applicable mandate was never reported — *not never paid* — and no
later submission backfills it. An award straddling FY2022 has its early
cash permanently missing from the record.

## Outlays in the tables you already have — and why not to trust them

The bulk transaction files carry one outlay field,
`total_outlayed_amount_for_overall_award` (canonical:
`award_total_outlayed`, surfaced on the award spine as `total_outlayed`).
It is a **lifetime-to-date snapshot** stamped onto every row at file
generation — there is no annual form — and it inherits every window
problem at once:

```{r era-table}
aw <- outlay_training$awards
d <- aw[linked == TRUE & oblig > 0 & !is.na(pop_end_fy) & pop_end_fy <= 2025]
d[, era := fifelse(first_fy >= 2022, "FY2022+", paste0("FY", first_fy))]
d[, .(awards = .N, median_outlay_over_oblig = round(median(outlay_total / oblig), 2)),
  keyby = era]
```

The gradient is the mandate, not the spending: FY2020 starts are missing
their pre-mandate cash, FY2021 starts part of it, FY2022+ starts almost
none. On top of the era problem, coverage varies by **agency practice** —
in the same completed, linked, mandate-era population, DoD reports File C
obligations but essentially no outlays (median ratio 0.00) and Commerce
0.24, while NSF, USDA and NASA sit at 1.00 and HHS at 0.91 — and by
**liquidation lag**: even clean-era awards read 0.90 a year after the
performance end and reach 1.00 only around three years. `us_reconcile()`
reports the ratio per award so the gap stays visible, but the lifetime
field is a floor with unknown slack, never an annual measure.

## Adding real annual outlays

Annual outlays exist in one place: the account-level File C records behind
`POST /api/v2/awards/funding/`, one row per federal account × object class
× program activity × reporting period. `us_fetch_outlays()` retrieves them
per award and `us_outlays_by_year()` collapses them to award × fiscal year
— applying the rule the raw field requires (`gross_outlay_amount` is
cumulative within each fiscal year, so the annual figure is the last
reported value per account cell, never the period sum).
`us_add_outlays()` attaches the result to a fiscal panel.

```{r, eval = FALSE}
p <- us_panel(extract, period = "fiscal")
p <- us_add_outlays(p)
p$panel[, .N, by = outlay_coverage]
```

### Limitations of that database and API

* **The era boundary** above: nothing before an award's mandate, ever.
* **Agency coverage**: for some portfolios (VA contracts, much of DoD)
  there is simply no cash to fetch, at any date.
* **Key and linkage challenges**: File C rows are linked to award
  identifiers by the submitting agency, and the linkage fails in both
  directions. Some awards return no File C records at all; others return
  records whose obligations do not reconcile with the award's own
  transaction ledger — about a fifth of fetched awards in this package's
  ground-truth pull failed a 10% linkage screen. The funding payload also
  carries **no recipient dimension**: outlays exist only at award grain,
  so multi-recipient awards cannot be split by organization from the cash
  side.
* **Operational limits**: one paged request per award, and sustained
  pulls beyond roughly a thousand requests can trip a temporary
  host-level block — large pulls run best throttled and cached.

`us_add_outlays()` encodes all of this per award as `outlay_coverage`
(`complete` / `truncated_pre_FY2022` / `unlinked` / `no_outlay_rows` /
`no_file_c` / `fetch_failed`), with `NA` — never a fake zero — where
measurement is absent.

## Where real outlays don't reach: imputation

For the majority of awards whose payment ledger is truncated, unlinked, or
absent, `us_impute_outlays()` estimates the cash calendar from the
obligations side. Two things matter about what it returns:

* The model is fitted only on awards whose payment ledgers are
  essentially **complete** (screened by completeness and shape — see
  `vignette("imputation-fitting")`), and it predicts the **typical
  payment schedule conditional on the obligations**: the shape of the
  cash calendar at the level completed awards actually deliver, about
  0.94 per obligated dollar.
* `reconcile = TRUE` re-imposes the accounting identity — every imputed
  series rescaled to sum exactly to net obligations (for in-progress
  awards, over the full projected life). That configuration makes imputed
  dollars tie out against the reconciled obligations ledger, which is
  what most analyses of government spending or organizational revenue
  want; the model then contributes what it genuinely knows, the timing.

`vignette("imputation")` walks the approach with drawn cases;
`vignette("reconciliation")` places both ledgers inside the package's
accounting identities.
