---
title: "Organizations, UEIs and the crosswalk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Organizations, UEIs and the crosswalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(datatable.print.trunc.cols = TRUE, width = 90)
```

```{r setup, message = FALSE}
library(usaspend)
library(data.table)
```

USAspending records money against a **UEI** — a SAM.gov registration. The
panel is about **organizations**. The two are not the same thing: one
organization can hold several registrations, a registration can be the
subsidiary of another, and the API itself blurs the line by returning a
parent's subsidiaries when you ask for the parent. This vignette covers how
the package maps UEIs to organizations — the **crosswalk** — and the kinds
of relationship that map has to represent.

In short:

- Every UEI whose own history was extracted is **in sample**. Everything
  else in the extract is kept for inspection but left out of the panel.
- `us_panel(org_map = )` assigns each in-sample UEI an `org_id`. Unlisted
  UEIs are their own organization, and subsidiaries follow their parent.
- `us_extract()` finds subsidiaries on its own. With `subsidiaries = FALSE`
  (the default) it records them and tells you it has not pulled them; with
  `subsidiaries = TRUE` it pulls their full histories and counts them as
  part of their parent.

## The default: one UEI, one organization

With no crosswalk, each requested UEI is its own organization and its
`org_id` is the UEI itself:

```{r}
ex <- us_sample_extract()
us_org_map(ex$meta$uei)
```

## Several registrations, one organization

Large nonprofits hold several SAM registrations, and USAspending splits
their awards across them: NYU has three UEIs, holding 2,687, 4,721 and 9
awards in the pilot. Request all of them and supply a crosswalk mapping
them to one `org_id` — an EIN, say. Every table downstream is then keyed on
the organization:

```{r}
# for illustration only: treat two of the sample's UEIs as one organization
om <- data.table(uei    = ex$meta$uei,
                 org_id = c("ORG-1", "ORG-1", "ORG-2"))
p <- suppressWarnings(us_panel(ex, org_map = om))
p$panel[, .(awards = uniqueN(award_key), net = sum(obligation_net)), by = org_id]
```

The crosswalk can only label data that was pulled. Rows for UEIs not in the
extract are ignored, with a message saying so. **Listing a UEI in `org_map`
does not add it to the sample; requesting it in `us_extract()` does.**

## Parents and subsidiaries

SAM records a parent for many registrations, and USAspending carries it on
every transaction as `recipient_parent_uei`. That matters because of how
the API filters.

### The API returns a parent's subsidiaries — partially

The bulk download filters on `recipient_search_text`, and **that filter
matches the parent UEI as well as the recipient UEI** (see
`?us_download_submit`). So a query for a parent also returns its
subsidiaries' transactions — but only those filed while the queried UEI was
*recorded* as their parent.

The measured case: querying RTI International (`JJHCMK4NT5N3`) returned
transactions of International Resources Group (`R29FEFR7P8H9`), which RTI
acquired in 2017. A download by award ID for one IRG contract found 14
transactions, all with IRG as recipient. The RTI extract held only the 2
filed with RTI as parent. The other 12, filed under IRG's earlier parents
L-3 and Engility, never came back. **A subsidiary arrives with its history
cut off at the date the parent relationship began.**

The effect is routine, not a corner case. In a 1,000-organization test pull,
every one of the 4,727 unrequested transactions came in this way: 18
subsidiary UEIs under 10 requested organizations. And "subsidiary" in SAM
covers more than acquisitions:

| kind | example from the test pull | parent-matched transactions |
|---|---|---|
| acquired firm | International Resources Group (acquired 2017) and Masimax Resources, under RTI | 317 |
| affiliated entities | CommonBond Housing, Skyline Tower of St. Paul LP and CB Meadow Village LLC, under CommonBond Communities | 95 |
| other registrations of the same institution | two Harding University UEIs recorded as children of the requested one | 4,006 |
| related operating units | four Ingenium Schools UEIs under a requested Ingenium UEI that has no awards of its own | 7 |

The Harding case shows how much can hang on this. The requested UEI holds
66 transactions of its own, and its two child registrations hold 4,006.
Leave the children out and most of the institution's federal money is
missing from its panel.

Whether each belongs in "the organization" is an analytic question, and the
package cannot answer it for you. What it does is make sure the question is
visible rather than answered by accident.

### What the extract records: the crosswalk

`us_extract()` records every UEI it saw in a crosswalk, `extract$org_map`:
requested UEIs, plus every unrequested recipient whose recorded parent is a
requested UEI. The sample has none, so the rest of this section simulates
one. It takes an award from the sample and re-attributes it to IRG's UEI,
with a requested UEI as the recorded parent — exactly what the API returns
for a subsidiary.

```{r}
KEY <- "ASST_NON_HDTRA12310001_097"
sim <- us_sample_extract()
tx  <- sim$transactions
k   <- tx$award_key == KEY
parent <- tx$recipient_uei[which(k)[1]]
tx$recipient_parent_uei[k] <- parent
tx$recipient_uei[k] <- "R29FEFR7P8H9"
sim$transactions <- tx

us_find_subsidiaries(sim)[, .(uei, relationship, parent_uei, root_uei,
                              extracted, n_parent_matched)]
```

The columns:

| column | meaning |
|---|---|
| `relationship` | `"requested"` or `"subsidiary"` |
| `parent_uei` | the parent recorded on the subsidiary's transactions |
| `root_uei` | the requested UEI at the top of the chain. A subsidiary's own subsidiaries, found when it is pulled, share its root |
| `org_id` | the organization it rolls up to. Until `us_panel()` applies your crosswalk, this is the root UEI |
| `extracted` | whether the UEI's **own** history is in the extract |
| `n_parent_matched` | transactions that arrived through the parent match |

`us_find_subsidiaries()` works on any extract, including ones made before
the crosswalk existed. `us_extract()` stores the same table as
`extract$org_map`.

### `subsidiaries = FALSE`: recorded, not pulled

This is the default. The subsidiaries go into the crosswalk with
`extracted = FALSE`, and when the extract finishes it tells you so:

```{r}
sim$org_map <- usaspend:::extract_crosswalk(sim)
usaspend:::report_subsidiaries(sim$org_map)   # what us_extract() prints
```

Their transactions stay in the extract but are **out of sample**:

```{r, warning = FALSE}
p <- us_panel(sim)
p$org_map[, .(uei, relationship, org_id, in_sample)]
KEY %in% p$panel$award_key            # left out of the panel
p$awards[award_key == KEY, in_sample] # kept in the spine, flagged
```

They are left out because the history is truncated. The award's reported
lifetime total counts every action, but the extract holds only the ones
filed under the requested parent. So `us_reconcile()` does not report the
gap as an unexplained break. It gives the award its own label, tested before
any other:

```{r, warning = FALSE}
r <- us_reconcile(p)
r[award_key == KEY, .(tx_sum, total_obligated, status)]
```

In the 1,000-organization test, 370 of 15,420 awards were out of sample,
and 41 of them had previously been counted as unexplained breaks ($232M of
gap).

### `subsidiaries = TRUE`: pulled in full, counted with the parent

```{r, eval = FALSE}
ex <- us_extract(uei, years = 2008:2025, subsidiaries = TRUE)

# or, on an extract you already have, without repeating the main pull:
ex <- us_add_subsidiaries(ex)
```

Each subsidiary UEI is then queried **in its own right**. A query on its
own UEI matches `recipient_uei`, so it returns the whole history, not only
the slice filed under the parent. The UEI joins the extracted set
(`ex$meta$uei`; the original request stays in `ex$meta$uei_requested`) and
is marked `extracted = TRUE`. A subsidiary's own pull can reveal
subsidiaries of its own; they are pulled in further rounds (`max_rounds`,
default 3). Rows that arrived through the parent match are not duplicated.
With `subawards = "out"` or `"both"`, the subsidiaries' awards get
pass-through fetched too.

What this recovers, measured on a live FY2016–2018 pull for RTI: the RTI
query returned 81 IRG transactions, all dated February 2017 or later. IRG's
own pull added 117 more, from October 2015 to February 2017, filed under
its previous parent. That is the history the parent match cannot see.

In the panel, an extracted subsidiary takes the `org_id` of the requested
UEI it rolls up to. You do not need to list it in your crosswalk:

```{r, warning = FALSE}
# simulate the second pull: mark the subsidiary extracted
full <- sim
full$org_map <- copy(sim$org_map)[uei == "R29FEFR7P8H9", extracted := TRUE]
full$meta$uei <- c(sim$meta$uei, "R29FEFR7P8H9")

om <- data.table(uei = sim$meta$uei, org_id = c("ORG-A", "ORG-B", "ORG-C"))
p  <- us_panel(full, org_map = om)
p$org_map[, .(uei, relationship, org_id, in_sample)]
unique(p$panel[award_key == KEY, org_id])
us_reconcile(p)[award_key == KEY, status]
```

An explicit row in `org_map` overrides this, for instance to keep a
subsidiary as an organization of its own:

```{r, warning = FALSE}
om2 <- rbind(om, data.table(uei = "R29FEFR7P8H9", org_id = "IRG"))
unique(us_panel(full, org_map = om2)$panel[award_key == KEY, org_id])
```

### Before the parent relationship began

A full history includes the years before the relationship existed. For
IRG, that is everything before RTI acquired it. With
`subsidiaries = TRUE` those years count toward the parent organization,
which is right if the question is "what does this organization's current
footprint look like over time" and wrong if it is "what was this
organization obligated in each year".

For the second question, drop a subsidiary's transactions filed under any
parent outside the crosswalk before building the panel:

```{r, eval = FALSE}
sub_uei <- ex$org_map[relationship == "subsidiary" & extracted == TRUE, uei]
tx <- ex$transactions
pre <- tx$recipient_uei %in% sub_uei &
  !tx$recipient_parent_uei %in% ex$org_map$uei
ex$transactions <- tx[!pre]
p <- us_panel(ex, org_map = om)
```

Expect those awards to fail reconciliation afterwards (`window_edge` if the
cut falls in the first year, otherwise `break`). You have deliberately cut
their histories, so the identity no longer holds.

### Requested UEIs whose parent was not requested

The match only runs downward: from a requested parent to its subsidiaries.
In the test pull, 42 requested UEIs recorded a parent that was *not*
requested. The package does not follow those upward. Pulling the parent
would bring in its whole portfolio, which the request did not ask for. To
see them:

```{r, eval = FALSE}
ex$transactions[recipient_uei %in% ex$meta$uei &
                  !is.na(recipient_parent_uei) &
                  recipient_parent_uei != recipient_uei &
                  !recipient_parent_uei %in% ex$meta$uei,
                .N, by = .(recipient_uei, recipient_parent_uei)]
```

## Relationships the crosswalk does not hold

Two other relationships look similar but live elsewhere, because they
belong to awards rather than to organizations:

- **An award held by several recipients.** NIH grants follow the principal
  investigator between institutions, so one award can carry transactions
  under several UEIs over its life. Both slices can be in sample, each
  belonging to its own organization. The panel grain is
  organization × award × year for this reason, and `us_reconcile()` labels
  these `multi_recipient`. An award counts as in sample if *any* of its
  transactions is on an extracted UEI.
- **Prime and subrecipient.** Money received as a subawardee is keyed
  through the subawardee UEI into `subawards_in`, not through the
  crosswalk. See `vignette("subawards")`.

## Summary

| relationship | how it enters | in sample? | `org_id` |
|---|---|---|---|
| requested UEI | `us_extract(uei)` | yes | from `org_map`, else the UEI |
| several registrations of one org | request all, map to one `org_id` | yes | shared `org_id` |
| subsidiary, `subsidiaries = FALSE` | found through the parent match | **no**: `out_of_sample`, history truncated | parent's, for reference |
| subsidiary, `subsidiaries = TRUE` or `us_add_subsidiaries()` | pulled on its own UEI | yes, full history | parent's, unless `org_map` lists it |
| `org_map` row for a UEI never pulled | — | no (ignored, with a message) | — |
| parent of a requested UEI | not followed | no | — |
| other recipients on a shared award | appear on the award's transactions | no, unless requested | — |
