---
title: "The panel: reading, rolling up, and deflating the output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The panel: reading, rolling up, and deflating the output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(datatable.print.trunc.cols = TRUE, width = 90)
```

Everything in this vignette runs offline, on the bundled sample extract — a
real three-UEI pull from USAspending kept verbatim in `inst/extdata/sample/`.

```{r setup, message = FALSE}
library(usaspend)
ex <- us_sample_extract()
p  <- us_panel(ex)
p
```

## What `us_panel()` returns

A `usaspend_panel` is a list of tables, not just the panel itself:

| element | grain | what it is |
|---|---|---|
| `panel` | org × award × year | **the deliverable** |
| `awards` | prime award | the spine: agency, type, CFDA/NAICS, lifetime totals |
| `transactions` | award action | the normalized ledger the panel was netted from |
| `subawards` | FSRS report line | normalized subawards, with direction |
| `subawards_in` | org × year | subaward dollars *received* — see below |
| `meta` | — | period, measure, de-obligation policy, build time |

```{r}
names(p)
p$panel[1:3, .(org_id, award_key, year, obligation_net, net_revenue)]
```

## How to read the dollar columns

Five rules cover most misreadings of the panel.

### 1. Obligations are commitments, not cash

`obligation_net` is the net federal obligation booked to the
organization-award-year: money the government *committed* by signing award
actions dated in that year. Federal award data carries no annual disbursement
figure at all, so an obligations panel is the closest thing to annual revenue
the source supports. On multi-year awards, obligations and cash differ by
years of timing.

### 2. Net revenue is obligations minus pass-through

An organization does not keep every dollar obligated to it. Money it is
obliged to pay onward to subrecipients — pass-through — is someone else's
revenue. The panel nets it out:

```
net_revenue = obligation_net − subaward_out_amount
```

But outbound subawards are only present if the extract fetched them
(`subawards = "out"` or `"both"` in `us_extract()` — they require a separate
API pass, one call per prime award). When they were not fetched, the panel
says so rather than letting a zero pass for a measured zero:

```{r}
unique(p$panel$flags)
p$meta$subawards_out_fetched
```

On this sample `net_revenue` equals `obligation_net` — true for any
recipient-filtered pull until the outbound pass runs.

### 3. Negatives are real; gross and net travel together

De-obligations — claw-backs, downward revisions, recoveries at closeout — are
ordinary federal award activity, and they arrive under every action-type
label, including `CONTINUATION`. The panel keeps them, which means an
org-award-year can legitimately net negative. It also always carries the
gross components, because $1M net from +$5M and −$4M is a different fact than
$1M from a single clean award:

```{r}
p$panel[obligation_negative < 0,
        .(award_key, year, obligation_positive, obligation_negative, obligation_net)]
```

### 4. Loans and student aid are separable, and loan face value is never revenue

Every row carries `award_family` (`grant`, `contract`, `direct_payment`,
`loan`, `other`, `idv`). Direct loans book their face value in
`loan_face_value` — a liability of the borrower, not income of the
organization — and their obligation is typically $0. Direct payments to
universities are mostly student aid (Pell, work-study) flowing *through* the
institution; filter on `award_family` to include or exclude them explicitly.

### 5. Inbound subawards are additional revenue — and they live in a separate table

A recipient-filtered pull returns subawards where your organization is the
**subawardee**: money received from another organization's federal award,
which appears nowhere in prime award data. It is real revenue, but it is
reported by the prime under the prime's award key, so it cannot sit on this
organization's award rows. It is aggregated per organization-year in
`subawards_in`:

```{r}
p$subawards_in
```

The sample makes the stakes concrete: compare prime obligations with inbound
subawards for the first organization —

```{r}
merge(
  p$panel[, .(prime_net = sum(obligation_net)), by = org_id],
  p$subawards_in[, .(inbound = sum(subaward_in_amount)), by = org_id],
  all.x = TRUE
)
```

`CFFMYPABYAG3` shows roughly \$220k of prime obligations against \$60M
received as a subrecipient. Judged on prime awards alone, this organization
barely exists as a federal recipient; nearly all of its federal money arrives
as pass-through from other primes' awards.

## Rolling up: `us_org_year()` and `us_rollup()`

Most analyses do not need award detail. `us_org_year()` collapses the panel
to organization × year (optionally by an extra panel column):

```{r}
us_org_year(p)[1:5]
us_org_year(p, by = "award_family")[1:5]
```

`us_rollup()` is the general aggregator: each dimension is a toggle, and it
folds the inbound table in so both directions of money end up in one row.

```{r}
us_rollup(p)                                 # one row per org, all years
us_rollup(p, org_id = FALSE, year = TRUE)[1:5]  # sector-wide trend
```

It adds one column the panel does not have:

```
total_net = obligation_net + subaward_in_amount − subaward_out_amount
```

— the fullest single measure of net federal dollars flowing to the group:
prime money kept, plus money received as a subrecipient. Use `net_revenue`
when the question is "what did this organization's *prime awards* net?", and
`total_net` when it is "how many federal dollars ended up here?".

`state = TRUE` groups by the award's recipient state (inbound flows, which
are organization-level, are assigned to the organization's dominant state):

```{r}
us_rollup(p, org_id = FALSE, state = TRUE)[, .(state, n_awards, obligation_net, total_net)]
```

## Constant dollars: `us_adjust_inflation()`

A panel spanning `r min(p$panel$year)`–`r max(p$panel$year)` mixes dollar
vintages. `us_adjust_inflation()` restates every dollar column into the
dollars of a target year, using the bundled CPI-U annual averages (or any
`year`/`index` table you supply — only ratios are used, so the base year
does not matter):

```{r}
tail(us_price_index(), 3)

trend   <- us_rollup(p, org_id = FALSE, year = TRUE)
trend25 <- us_adjust_inflation(trend, target_year = 2025)

cbind(trend[year %in% 2008:2010, .(year, nominal = obligation_net)],
      trend25[year %in% 2008:2010, .(real_2025 = obligation_net)])
```

The adjustment records what it did:

```{r}
str(attr(trend25, "usaspend_inflation"))
```

Order matters: adjust **before** collapsing across years. A rollup that has
already summed 2008 dollars with 2024 dollars has no `year` column left to
deflate by, and the function refuses rather than guessing:

```{r, error = TRUE}
us_adjust_inflation(us_rollup(p))   # collapsed across years -- refuses
```

You can also pass the whole `usaspend_panel`, which adjusts `panel` and
`subawards_in` together and notes the target year in `meta`.

## Checking a build

Every rule that produced these numbers is a choice, and choices need checks.
`us_reconcile()` tests each award's summed transactions against the lifetime
total USAspending computes independently, and `us_audit()` runs the
structural invariants (no duplicate transactions, unique grain, no award
attributed to two organizations). Run both on every rebuild.
