---
title: "Reconciliation: accounting identities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reconciliation: accounting identities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
p <- us_panel(us_sample_extract())
```

The goal of this package is to start from a **complex transaction ledger**
— tens of thousands of modification rows carrying corrections, restated
duplicates, zero-dollar administrative actions, claw-backs, and ceilings
that are not money — and convert it into an **accounting ledger**: one row
per organization × award × year whose dollars add up under stated rules.
"Under stated rules" is the whole game. Every rule is a choice, and a
choice that cannot be checked against an independent figure is an
assumption. This vignette lays out the identities the package imposes,
then shows how they are tested.

## The identities

### What adds, what subtracts, what has no effect

The money column is `federal_action_obligation`, and **the sign is the
truth** — the action-type code describes intent, never direction
(a CONTINUATION can carry negative dollars; 6,344 did in the pilot).

* **Adds to net**: any positive obligation — originations (assistance `A`,
  contract base awards), funded continuations (assistance `B`, contract
  option exercises `G`), positive revisions and funding-only actions
  (contract `C`). A *program expansion* looks like this: new positive
  obligations, usually alongside a period extension.
* **Subtracts from net**: any negative obligation — de-obligations and
  claw-backs, arriving under every action class (terminations, negative
  continuations, negative revisions, closeout adjustments). The
  `deobligation_policy` argument of `us_panel()` chooses *when* they
  subtract: the year posted (`as_posted`), restated LIFO against the
  award's earlier positive years (`restate`), or measured-and-discarded
  (`drop`).
* **No effect on net**: zero-dollar actions — 20–30% of all rows. These
  carry the administrative record: period-of-performance changes,
  closeouts, vendor and transfer updates. A *timeline-only extension* is
  the canonical case: the end date moves, the money does not, net is
  untouched. (A non-zero amount on an administrative action is flagged
  `money_on_admin_action` rather than silently trusted.)
* **Never in net at all**: loan face value (the borrower's liability),
  loan subsidy cost, non-federal cost share, and every ceiling —
  `potential_value`, IDV vehicle totals. The pilot holds $115bn of IDV
  ceiling against $0.75bn actually obligated on those vehicles.

### The award window, and whether it binds

The **period of performance** (`pop_start_date` → `pop_end_date`) is the
award's declared window. It is *descriptive, not binding*: transactions
legitimately post after the end date (corrections, closeout adjustments),
and the normalizer only flags actions landing more than a year past it
(`action_after_pop`) rather than rejecting them.

The window also **moves**. The data signature of a period change is a
`pop_end_date` that differs across an award's transaction sequence — the
first reported value is the original plan, the last is the final window,
and `us_outlay_features()` classifies the difference: an extension past 90
days with material new money is `extension_funded` (an expansion), without
new money `extension_timeline` (a no-cost extension), and a shortened
window usually rides along with a `reduced` award's de-obligations.

**Closeout is a third clock.** The closeout action (contract code `K`,
assistance `D` "Adjustment to Completed Award") follows neither the
performance end nor the last payment: measured on File C ground truth,
cash is still arriving a year after the performance end (outlay/obligation
ratio 0.90 at one year, 1.00 only at three), and de-obligations of genuine
underspend post later still. Performance end → final drawdowns → audit and
closeout is a sequence spanning years, and each stage leaves rows.

### Two year definitions

Every transaction carries both `action_year` (calendar) and
`action_fiscal_year` (federal fiscal: October–September), and `us_panel()`
builds on either (`period =`). The choice is an accounting decision:
October–December actions belong to different years under the two
definitions, so panels built on different bases do not reconcile
year-by-year, only in lifetime totals. Match the basis to the linked data
— calendar for IRS Form 990 and most organizational outcomes, fiscal for
anything meeting the federal ledgers (File C outlays exist *only* in
fiscal periods, which is why `us_add_outlays()` and
`us_add_imputed_outlays()` require a fiscal panel).

### Pass-through: the prime's ledger and the sub's

A prime award's obligations are the organization's **gross** federal
revenue. What it is obliged to pay onward to subrecipients is measured
separately (`subaward_out_amount`), and the identity

> `net_revenue = obligation_net − subaward_out_amount`

defines the **retained** column: what the prime keeps after pass-through.
The same dollar deliberately appears twice in the system — once in the
prime's gross, once as the subrecipient's inbound revenue
(`subaward_in_amount`, kept in the separate org-year `subawards_in` table)
— because both ledgers are true at their own level. What the package never
does is add them in one total.

### Why a subaward row will not reconcile

Subaward data is **not a transaction ledger**. FSRS rows are *report
lines*: one line per subaward as last restated, no modification history,
no de-obligations, reported only above the $30k threshold, and lagging the
obligations they draw on (13% of pilot pass-through dollars posted in
years with no prime activity). There is no lifetime total to test a
subaward sum against, so subaward figures are treated as **floors**
carried alongside the reconciled prime ledger — never as a ledger that
must balance.

## Testing the identities

The tests check two distinct assumptions:

1. **The database and the retrieval are consistent** — the extract holds
   what USAspending holds, without duplication or loss along the download
   path.
2. **The transaction-to-accounting transformation gets the math right** —
   de-duplication, delete handling, sign netting, and year assignment
   reproduce the totals they should.

A single external figure tests both at once: USAspending publishes each
award's **lifetime obligation total, computed by a different system than
the transactions**. If the extract is complete and the netting is correct,
summed transactions return to that total. `us_reconcile()` runs the
identity per award and labels every exception:

```{r, warning = FALSE}
r <- us_reconcile(p)
r[, .(award_key = substr(award_key, 1, 34), total_obligated, tx_sum, gap, status)]
```

### Reading the labels

The sample reconciles 15 of 18 exactly, and the exceptions carry the two
lessons:

- **`no_reported_total`** — nothing to check against; a coverage fact, not
  a failure.
- **`break`** — the claw-back award shows `tx_sum = −$310,999` against a
  $1.39M lifetime total because the extract's window holds only the
  reversal. The identity *should* fail here: **a UEI-filtered extract is
  an organization-eye view, not an award-eye view**, and the panel
  measures what this organization was obligated in these years — which
  legitimately differs from whole-life totals whenever the award's history
  extends beyond the window (`window_edge`, `recent_open`) or the award
  moved between institutions (`multi_recipient`).

### The same test at scale

On the full 50-org pilot (61,738 awards), after normalization:

| status | awards | explanation |
|---|---|---|
| **ok — exact to the dollar** | 44,680 (72%) | both assumptions hold |
| no reported total | 4,955 | nothing to check against |
| recent/open award | 4,133 | corrections keep arriving after the pull (API-audited) |
| window edge | 1,564 | pre-2008 history sits below the search floor |
| multi-recipient | 129 | the award's other slice belongs to another organization |
| unexplained | ~10% | API audit: mostly the same mechanisms mid-window, dominated by NIH awards following a PI between institutions |

72% exact — with every exception *labeled* — is a strong result for a
UEI-filtered extract, and the labeling is the point: a falling "ok" share
on a rebuild is the earliest sign of a broken rule, and an unlabeled break
is a real question, not noise. De-duplication is what makes the number
possible: before the duplicate rule (`vignette("accounting")`), 5,754
additional pilot awards failed the identity.

### Structural invariants — `us_audit()`

Checks needing no external reference:

```{r, warning = FALSE}
us_audit(p)
```

No transaction key twice, no award attributed to two organizations, one
row per org-award-year — plus the counters worth watching over time
(negative years, zero-dollar share, anomaly flags) and a warning whenever
outbound subawards were never fetched, so `net_revenue = obligation_net`
cannot pass silently as a netted figure.

## Obligations versus outlays

Everything above concerns the **budget ledger** — obligations, the
promises. The **payment ledger** — outlays, the cash — is a separate
system with its own identity: money obligated is eventually either paid
out or de-obligated, so lifetime outlays converge to lifetime net
obligations at closeout. The data confirms the identity where measurement
allows: awards with full reporting coverage and two-plus years to
liquidate reach a median outlay/obligation ratio of exactly 1.00.

But the payment ledger as *published* cannot be trusted the way the
budget ledger can (`vignette("obligations-outlays")` has the full story):
account-level outlay reporting was optional before the FY2022 monthly
mandate, whole agencies still report obligations without outlays (DoD
median ratio 0.00, Commerce 0.24, against NSF/USDA/NASA at 1.00), and a
fifth of fetched awards have File C records that cannot be linked back to
their own transaction ledger. `us_fetch_outlays()` /`us_add_outlays()`
retrieve what exists and grade each award's coverage rather than letting
gaps read as zeros.

Where real outlays don't reach, `us_impute_outlays()` fills in — and it is
important to be precise about what it returns. The model is fitted on
awards whose payment ledgers are essentially complete (screened by
completeness and shape), and it predicts the **typical payment schedule
conditional on the obligations**: the shape of the cash calendar, at the
level completed awards actually deliver (~0.94 per obligated dollar), not
the accounting identity. Passing `reconcile = TRUE` re-imposes the
identity — each award's imputed series is rescaled to sum exactly to its
net obligations (for an in-progress award, over its full projected life
rather than the years observed so far). That is the configuration most
users modeling government spending or organizational revenue want: the
dollars tie out to the reconciled obligations ledger, and the model
contributes what it genuinely knows, the timing.

## Practice

Run `us_reconcile()` and `us_audit()` on every rebuild and diff against
the previous run, so a change in the data or in a rule shows up as a
change in the audit rather than as a quietly different number.
