---
title: "Reconciliation: why 72% is a good number"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reconciliation: why 72% is a good number}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
p <- us_panel(us_sample_extract())
```

Every rule in this package is a choice, and a choice that cannot be checked
against an independent figure is an assumption. USAspending happens to
provide the independent figure: **award-lifetime totals computed by a
different system than the transactions.** `us_reconcile()` tests, for every
award, whether the summed transaction obligations come back to that total.

```{r, warning = FALSE}
r <- us_reconcile(p)
r[, .(award_key = substr(award_key, 1, 34), total_obligated, tx_sum, gap, status)]
```

## Reading a break

The sample reconciles 15 of 18 exactly. The classified exceptions teach the
two lessons that matter:

- **`no_reported_total`** — some awards simply publish no lifetime total to
  check against. Not a failure; a coverage fact.
- **`break`** — the interesting one. Here the claw-back award shows
  `tx_sum = −$310,999` against a lifetime total of $1.39M, because this
  extract's window holds only the reversal — the original obligations
  predate it. The identity *should* fail: **a UEI-filtered extract is an
  organization-eye view, not an award-eye view.** The panel measures what
  this organization was obligated in these years, which legitimately differs
  from the award's whole-life total whenever the award's history extends
  beyond the window or the award moved between institutions.

## The same check at scale

On the full 50-org pilot (61,738 awards), after normalization:

| status | awards | explanation |
|---|---|---|
| **ok — exact to the dollar** | 44,680 (72%) | the identity holds |
| no reported total | 4,955 | nothing to check against |
| recent/open award | 4,133 | corrections keep arriving after the pull (API-audited) |
| window edge | 1,564 | pre-2008 history sits below the search floor |
| multi-recipient | 129 | the award's other slice belongs to another organization |
| unexplained | ~10% | API audit says: mostly the same mechanisms mid-window, dominated by NIH awards following a PI between institutions |

72% exact — with every exception *classified* — is a strong result for a
UEI-filtered extract, and the classification is the point: a falling "ok"
share on a rebuild is the earliest sign of a broken rule, and an
unclassified break is a real question, not noise.

De-duplication is what makes the number possible at all: before the
duplicate rule (`vignette("accounting")`), 5,754 additional pilot awards
failed the identity.

## The other checks — `us_audit()`

Structural invariants, run on every rebuild:

```{r, warning = FALSE}
us_audit(p)
```

No transaction key twice, no award attributed to two organizations, one row
per org-award-year — plus the descriptive counters worth watching over time
(negative years, zero-dollar share, anomaly flags) and a warning whenever
outbound subawards were never fetched, so `net_revenue = obligation_net`
cannot pass silently as a netted figure.

## The outlay gap

`us_reconcile()` also reports `outlay_ratio` — lifetime outlays over
lifetime obligations — where USAspending provides outlays (mostly
post-FY2017 awards). This is the promise-versus-payment gap: the number that
tells a reader how far an obligations panel is from a cash panel
(`vignette("data-model")` explains why no annual cash figure exists).

## Practice

Run both checks on every rebuild and diff against the previous run, so a
change in the data or in a rule shows up as a change in the audit rather
than as a quietly different number.
