---
title: "Subawards: pass-through, direction, and what FSRS cannot see"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subawards: pass-through, direction, and what FSRS cannot see}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
ex <- us_sample_extract()
```

## What a subaward row is — and is not

A subaward row is an **FSRS report**: the prime awardee naming a first-tier
subrecipient, required once a subaward crosses the **$30,000** threshold.
Both parties carry UEIs, which is what makes direction recoverable at all.

It is **not** the federal → state → local funding chain in general. When CMS
pays a state Medicaid agency and the state pays a hospital, USAspending sees
only the federal → state prime award; the state's onward payment is not an
FSRS record, and Medicaid provider payments never appear at all. This is the
single most common misreading of the data.

The pattern *does* surface where the pass-through entity is itself a federal
prime grantee that files FSRS — and the sample catches it live. Its largest
inbound subaward is FEMA disaster money reaching Kaiser **through the
California Governor's Office of Emergency Services**, a state agency that is
the prime grantee:

```{r, warning = FALSE}
om <- us_org_map(ex$meta$uei)
sb <- us_normalize_subawards(ex$subawards, org_uei = om$uei)
sb[order(-subaward_amount)][1, .(prime_name, subawardee_uei, subaward_amount,
                                 subaward_action_date)]
```

So state-conduit flows appear exactly when the state is a prime grantee that
reports — a meaningful but **incomplete** slice of intergovernmental
transfers. No analysis may treat observed subawards as the population of
pass-through.

## Direction is the whole problem

A recipient-filtered bulk pull matches on the **subawardee**, so it returns
only rows where your organization *receives*. Measured twice: this sample's
32 rows are all inbound; the full pilot's 39,243 rows split 29,319 inbound,
2,497 internal (both sides pilot orgs — NYU → NYU School of Medicine and the
like), and **zero outbound**.

```{r}
sb[, .N, by = direction]
```

The four directions and what the panel does with each:

| direction | meaning | treatment |
|---|---|---|
| `out` — org is prime | pass-through obligation | **subtracted**: `net_revenue = obligation_net − subaward_out` |
| `in` — org is subawardee | revenue invisible in prime data | added as its own org-year table |
| `internal` — both sides yours | transfers between an org's own UEIs | excluded from both flows ($551m in the pilot) |
| `unrelated` | multi-recipient award now held elsewhere | excluded |

Direction cannot be read off a row alone — it needs the organization's UEI
set, which is why `us_normalize_subawards()` takes `org_uei` and
`us_subaward_by_year()` errors rather than guessing.

**The asymmetry to remember:** inbound arrives free with your extract;
outbound costs a separate fetch by prime award; and only outbound changes
`net_revenue`.

## Fetching outbound at scale

`us_extract(subawards = "out")` walks the extracted award keys through
`us_fetch_subawards_batch()` — batched ~400 prime award ids per request.
The pilot's 61,738 awards ran in 147 batches × 2 families, ~65 minutes,
zero failures. Two traps it absorbs, both hit live: mixing assistance and
contract type codes in one request makes the results endpoint fail *silently*
(counts say subawards exist; results return none), so it runs one pass per
family; and `award_ids` matches bare FAINs/PIIDs, which agencies reuse —
four of five raw rows returned were strays belonging to other awards, so
results are filtered back to the exact generated prime-award key. Without
that filter, pass-through inflates roughly fivefold.

## Tabulating so the numbers reconcile

- **De-duplicate restatements.** FSRS reports are restated month to month;
  19% of pilot bulk rows were the same subaward re-reported. Keep the latest
  report per composite key.
- **Book on `subaward_action_date`**, not the report date — reports lag,
  sometimes across fiscal years.
- **Attach with `us_panel(fill_gaps = TRUE)`.** FSRS reports trail the
  obligations they draw on: 13% of pilot outbound dollars ($1.7bn) were
  dated in org-award-years with *no prime activity at all*. Without
  synthetic zero-obligation rows those dollars silently drop.
- **Never attach one subaward to two organizations** — multi-recipient
  awards would double-count it; the panel attaches once, to the dominant
  org-row.

Reconciliation target, asserted in the package tests: the panel's
`subaward_out_amount` sums to exactly the dollars fetched.

## What it amounts to

Pilot magnitudes: **$12.6bn outbound and $14.0bn inbound against $118.5bn
of net prime obligations** — net revenue $105.9bn after pass-through.
Concentration is extreme (the top four orgs carry over half the outbound
dollars; one research institute passes through 45% of everything it
receives), and an organization funded purely as a subrecipient looks like a
non-recipient in prime data.

## Known limits — carry these next to any subaward number

First-tier only (no sub-sub-awards); under-$30k censored, so pass-through is
a lower bound everywhere; reporting compliance is uneven, so "zero
subawards" and "never reported" are indistinguishable at the row level (keep
counts alongside dollars); the subawardee UEI is often null for government
subawardees; and FSRS amounts are commitments, not disbursements.
