---
title: "Get started: UEIs in, panel out"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started: UEIs in, panel out}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

usaspend turns a list of SAM.gov Unique Entity Identifiers (UEIs) into a
clean **organization × award × year** panel of net federal award activity
from USAspending.gov. The whole path is three calls:

```{r, eval = FALSE}
library(usaspend)
ex <- us_extract(c("CFFMYPABYAG3", "FG8QB99NF8K3"),   # your UEIs
                 years = 2008:2025, subawards = "both")
p  <- us_panel(ex)                                     # normalize + net
us_rollup(p, year = TRUE)                              # analyze
```

Everything below runs offline, on the bundled sample extract — a real
three-UEI pull kept verbatim.

```{r setup, message = FALSE, warning = FALSE}
library(usaspend)
ex <- us_sample_extract()
p  <- us_panel(ex)
p
```

## The one idea everything else depends on

**A federal award is not a payment. It is a running ledger of promises,
written in modifications.** Here is a real award from the sample — eight
actions across four calendar years, three of them zero-dollar:

```{r}
p$transactions[award_key == "ASST_NON_HDTRA12310001_097",
  .(modification_number, action_date, action_type_label,
    federal_action_obligation)][order(action_date)]
```

The panel books each action's dollars to the year it happened:

```{r}
p$panel[award_key == "ASST_NON_HDTRA12310001_097",
  .(year, n_transactions, obligation_net)]
```

And the check that the whole pipeline is right: the years must sum back to
the award's reported lifetime total, which USAspending computes in a
different system than the transactions —

```{r}
sum(p$panel[award_key == "ASST_NON_HDTRA12310001_097"]$obligation_net)
p$awards[award_key == "ASST_NON_HDTRA12310001_097"]$award_total_obligated
```

That identity, checked across every award by `us_reconcile()`, is what makes
the panel trustworthy rather than merely plausible. A reader who has
internalized the ledger idea — actions, signs, years, and a lifetime total
to reconcile against — can predict most of the package's design.

## Where everything is explained

| vignette | what it covers |
|---|---|
| `vignette("data-model")` | how USAspending is structured: transactions, awards, subawards — and why the panel can only be built from transactions |
| `vignette("award-types")` | grants, contracts, IDVs, loans, direct payments — and which of them are revenue |
| `vignette("acquisition")` | the two ways to fetch: API jobs vs annual archives, measured |
| `vignette("accounting")` | the reconciliation rules: de-duplication, corrections, de-obligations, and which year a claw-back belongs to |
| `vignette("subawards")` | pass-through and subrecipient revenue: what FSRS sees, and direction |
| `vignette("panel")` | reading the output; rollups; inflation adjustment |
| `vignette("reconciliation")` | the checks: the lifetime identity, and how to read a break |
| `vignette("structure")` | package architecture and the function-call workflow |
